---
title: "CMDA-4654  \n Group Exercise 1B"
author: "Sophia Gavasheli and Jacob Carlson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    highlight: kate
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---
  

```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here
library(readr)
library(dplyr)
library(ggplot2)
library(e1071)
library(caret)
library(FNN)

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE, # Turn off warnings
                      R.options = list(width = 100))

library(here)
setwd(here()) # sets path to root of RProject

```

\clearpage



# 1. Reintroduction
We are Sophia Gavasheli and Jacob Carlson, and we our group 17. Our chosen domain is the sports category.

# 2. Executive Summary and Research Questions 

## KNN Regression
We are using a KNN regression model to predict the runs a team scored in any given game. Runs will be predicted by the following variables.

| Variable   | Description                                           |
|------------|-------------------------------------------------------|
| `AB`       | At Bats                                               |
| `1b`       | Singles                                               |
| `2B`       | Doubles                                               |
| `3B`       | Triples                                               |
| `HR`       | Home Runs                                             |
| `SB`       | Stolen Bases                                          |
| `CS`       | Caught Stealing                                       |
| `BB`       | Base on Balls                                         |
| `SO`       | Strikeouts                                            |
| `HBP`      | Hit by Pitch                                          |
| `SH`       | Sacrifice Hits                                        |
| `SF`       | Sacrifice Flies                                       |

The data frame that these variables were taken from was not used in part A of our exercise. Using runs as a metric was an explored idea in part A, but was not used in our final report, so we decided to further explore it here, as it was a simple an obvious choice for regression.

## KNN Classification and Naive Bayes
For our KNN Classification and Naive Bayes models, we chose to predict a players position using their offensive stats.

| Variable   | Description                                           |
|------------|-------------------------------------------------------|
| `BA`       | Batting Average                                       |
| `HRA`      | Home Runs Average                                     |
| `BBA`      | Base on Balls Average                                 |
| `SOA`      | Strikeout Average                                     |

We standardized this data to ensure KNN and Naive Bayes were using the same inputs. We actually spent a lot of our part A analysis on position metric, so exploring further with the two models was a natural part B. 

In part A, we used the data frames `fieldingpost.csv` and `batting.csv`. Fielding-post only included post season stats(playoffs), and wasn't big enough for our models, so we switched to `fielding.csv`, which contains the entire seasons. We also used `Teams.csv` for our KNN regression, as it contained all of the data used in our model with minimal cleaning and merging between data sets.

The data is linked [here](https://sabr.app.box.com/s/y1prhc795jk8zvmelfd3jq7tl389y6cd). We downloaded `Fielding.csv`, `batting.csv`, and `Teams.csv`.

```{r,echo = FALSE}
pos     <- read.csv("Fielding.csv")
batting <- read.csv("batting.csv")
team    <- read.csv("Teams.csv")
# Getting highest position played for players
primary_pos <- pos %>%
  group_by(playerID, yearID, POS) %>%
  summarise(G = sum(G), .groups = "drop") %>%
  group_by(playerID, yearID) %>%
  slice_max(G, n = 1, with_ties = FALSE) %>%
  select(playerID, yearID, POS)

# Merge Position into Player and Year to condense data
stats <- merge(batting, primary_pos,
               by = c("playerID", "yearID"), all.x = TRUE)

# Keeps only American and National League
stats_clean <- stats[stats$lgID %in% c("AL", "NL"), ]
stats_clean <- stats_clean[!is.na(stats_clean$POS), ]


#####################################################
####### REGRESSION !!!!!!!!!
## Separates by team for regression prediction. 

team_clean <- team[team$lgID %in% c("AL","NL"),] # Keeps only AL and NL
team_clean$X1B <- team_clean$H - team_clean$X2B - team_clean$X3B - team_clean$HR # Creates X1B
team_clean$HBPxBB <- team_clean$BB + team_clean$HBP # Creates HBPxBB


# Takes only the necessary data
team_clean <- team_clean[,c("yearID", "teamID","G", "R", "X1B", "X2B", "X3B", 
                            "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")] 

#Divides stats by games to establish per game base
stats <- c("R", "X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF")
team_clean[stats] <- team_clean[stats] / team_clean$G

team_clean <- team_clean[!is.na(team_clean$SF), ]# Removes NaN values from SF
team_clean <- team_clean[!is.na(team_clean$HBPxBB), ]# Removes NaN values from HBPxBB

# List of predictors for training
num_vars <- c("X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB") 


###################################################

# CLASSIFICATION AND BAYES!!!!!!!!!!!!!
# Keeps only At Bats > 100
pos_data <- stats_clean[stats_clean$AB >= 100, ] 

# Changes The infield positions into a common infield tab
for(i in 1:length(pos_data$POS)){
  if(pos_data$POS[i] == "1B" || 
     pos_data$POS[i] == "2B" ||
     pos_data$POS[i] == "3B" ||
     pos_data$POS[i] == "SS" ){
    pos_data$POS[i] = "IF"
  }
}
# For easier referral
position_names <- c(
  "IF" = "Infield",
  "C"  = "Catcher",
  "OF" = "Outfield",
  "P"  = "Pitcher"
)

#Factors Positions into levels
pos_data$Position <- factor(position_names[pos_data$POS])

#Keep only necessary data
pos_data <- pos_data[, c("playerID", "yearID", "Position",
                         "H", "HR", "BB", "SO", "SB", "AB")]

# creates scaled pos_data for Classification comparison and naive Bayes
pos_scale <- pos_data

pos_scale$BA <- pos_data$H  / pos_data$AB
pos_scale$HRA <- pos_data$HR / pos_data$AB
pos_scale$BBA <- pos_data$BB / pos_data$AB
pos_scale$SOA <- pos_data$SO / pos_data$AB

# Factors Pos_Scaled
pos_scale$Position <- factor(position_names[pos_data$Position])

#Adds scaled variables into pos_scaled
pos_scale <- pos_scale[, c("playerID", "yearID", "Position",
                                  "BA", "HRA", "BBA", "SOA")]
```
\clearpage

# 3. Methodology & Modeling Setup

## Training Split
We kept with the 70/30 training split that was found in our notes. The data sets were very large, and we preferred to keep the majority in our training.

## Data Reprocessing
### KNN Regression
For our KNN regression, `Teams.csv` includes a great many more predictors than what was used.
We build the data frame `team_clean`, which keeps only the American and National leagues, removes NaN rows, and takes in only our predictors, TeamId, yearID, and our prediction variable Runs. We also combined the HBP and BB predictors into one, as they provide the same outcome in a game. We also divided all of our predictors by games to establish the per game basis for our prediction.

We standardized our data set under the basic z-score scaling. As mentioned above, the predictor `HBPxBB` was a sum of two predictors, and was unfairly weighted against our other data. Furthermore, given the nature of the game, some of our predictors had a reported 0 weight on our distance, so standardizing was necessary.

#### Raw table
```{r, echo = FALSE}
#creates table for unscaled distances
d2 <-(team_clean[1, num_vars] - team_clean[2, num_vars])^2
round(rbind(squared_diff = unlist(d2), percent = 100*unlist(d2) / sum(d2)),2)
```

#### Scaled Table
```{r, echo = FALSE}
# table for scaled distances
z <- scale(team_clean[, num_vars])
d2z <- (z[1,] - z[2,])^2
round(rbind(squared_diff = d2z, percent = 100*d2z/sum(d2z)),2)
```

### KNN Classification and Naive Bayes
For our KNN classification and Naive Bayes models, we first created `primary_pos`, which read through the player's data, found there most played position, and changed their other reported positions to the highest one. We then combined `primary_pos`with batting, based on their playerId and the yearID, so that we could compare the positions to their offensive stats. We also took only the American and National league players to follow our analyis in part A. 

We wanted to limit classification responses, so we combined the infield positions (1B,2B,3B, SS) into an `Infield` category. 

In baseball, different positions get differing amounts of playtime. Furthermore, because we combined four possible positions into one, our data was even more skewed, and in the case of KNN classification, our distances would be poorly weighted. We therefore decided to scale all of our predictors based on At bats, providing the predictors described above(BA, HRA, BBA, SOA). While the standardizing would have less affect on our Naive Bayes model, since our goal is to compare it with our classification, we chose to scale the predictors before running either model.

All of our predictors were numerical, so there was no need to handle categorical leveling or do any smoothing to our data.

\clearpage

# 4. Analysis, Tuning & Results

##  KNN Regression

###  K Tuning with Cross Validation
```{r,echo = FALSE}

#splits training/testing 70/30
set.seed(4654)
index <- sample(1:nrow(team_clean), round(nrow(team_clean) * 0.7))
training_df <- scale(team_clean[index, num_vars])
testing_df  <- scale(team_clean[-index, num_vars],
                     center = attr(training_df, "scaled:center"),
                     scale  = attr(training_df, "scaled:scale"))

train_pred <- team_clean$R[index]
test_pred <- team_clean$R[-index]

# Cross Validation for best K
set.seed(4654)
folds <- sample(rep(1:10, length.out = nrow(training_df)))

cv_rmse <- function(k) {
  sq_err <- unlist(lapply(1:10, function(f) {
    tr_x <- training_df[folds != f, ]
    te_x <- training_df[folds == f, ]
    tr_y <- train_pred[folds != f]
    te_y <- train_pred[folds == f]
    pred <- knn.reg(train = tr_x, test = te_x, y = tr_y, k = k)$pred
    (te_y - pred)^2
  }))
  sqrt(mean(sq_err))
}

k_cv <- 1:80
cv_results <- data.frame(k = k_cv, RMSE = sapply(k_cv, cv_rmse))
k_best <- cv_results$k[which.min(cv_results$RMSE)]

#Training knn for values
train_rmse <- function(k) {
  pred <- knn.reg(train = training_df, test = training_df, y = train_pred, k = k)$pred
  sqrt(mean((train_pred - pred)^2))
}
cv_results$Train_RMSE <- sapply(cv_results$k, train_rmse)

# plot showing best k and training rmse results
ggplot(cv_results, aes(x = k)) +
  geom_line(aes(y = RMSE, color = "10-fold CV"), linewidth = 1) +
  geom_line(aes(y = Train_RMSE, color = "Training"), linewidth = 1) +
  geom_point(data = cv_results[cv_results$k == k_best, ],
             aes(y = RMSE), color = "red", size = 4) +
  scale_color_manual(values = c("10-fold CV" = "#8B2346", "Training" = "gray50")) +
  labs(title = paste0("Cross-validated choice:  k = ", k_best),
       x = "k", y = "RMSE", color = NULL) +
  theme_bw(base_size = 14)


top_k <- head(cv_results[order(cv_results$RMSE), ], 5)
top_k$RMSE   <- round(top_k$RMSE, 3)
rownames(top_k) <- NULL
top_k # plot showing top 5 ks and their RMSE
```

\clearpage 

###  Final predictions compared to null on k = 8

```{r,echo=FALSE}
final_pred <- knn.reg(train = training_df, test = testing_df,
                      y = train_pred, k = k_best)$pred

rmse      <- function(a, b) sqrt(mean((a - b)^2))
mae       <- function(a, b) mean(abs(a - b))
r_squared <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)

null_prediction <- rep(mean(train_pred), length(test_pred))

results <- data.frame(
  Model = c("Null (training mean)", paste0("KNN (k = ", k_best, ")")),
  RMSE  = c(rmse(test_pred, null_prediction), rmse(test_pred, final_pred)),
  MAE   = c(mae(test_pred, null_prediction),  mae(test_pred, final_pred)),
  R2    = c(r_squared(test_pred, null_prediction), r_squared(test_pred, final_pred))
)
results[, -1] <- round(results[, -1], 3)
results
```
#### RMSE

Our null model was off by about half of a run every game when predicting the training average. Our k = 8 was off by about a quarter of a run every game, which is about a 50% reduction compared to our null model.

#### MAE
Our null models' absolute average miss is a little under half a run every game. I k=8 model was off a little under half that, also being a little more than a 50% reduction compared to the null.

#### $R^2$
As expected, our null model takes 0% of the predictors into account for its predictions. However, our k = 8 model takes close to 81%, signifying that the 10 predictors are a strong representation of the runs scored.

\clearpage

##  KNN Classification



##  Naive Bayes
```{r,echo = FALSE}
H <- table(pos_scale$Position)# Establish count
prior <- prop.table(H)# establish prior
rbind(Count = H, Prior = round(prior, 4))# binds count and prior to one table

```
A basic Count/Prior table has the catcher and outfield positions being the most populated.
```{r,echo = FALSE}
# Splits train/test 70/30
set.seed(4654)
idx <- unlist(lapply(split(1:nrow(pos_scale), pos_scale$Position),
                     function(i) sample(i, round(0.7 * length(i)))))
trainNB <- pos_scale[idx,]
testNB <- pos_scale[-idx,]
# gives categorical likelihood tables
fitNB <- naiveBayes(Position ~ BA + HRA + BBA + SOA, data = trainNB) 
fitNB$tables 
```

```{r,echo = FALSE}

# class prediction and its posterior
mapNB <- predict(fitNB, newdata = testNB, type = "class") 
postNB <- predict(fitNB, newdata = testNB, type = "raw")

#data frame of predictions and posteriors
pred_dfNB <- data.frame(
  predicted = mapNB,
  posterior   = round(apply(postNB, 1, max), 4)
)
head(pred_dfNB, 10)
# Prediction matrix
tabNB <- table(predicted = predict(fitNB, newdata = testNB), actual = testNB$Position)
tabNB
```
I have produced the first 10 hard predictions and soft posteriors. Below that is the accompanying prediction matrix.

```{r,echo = FALSE}
# Confusion matrix 
cmNB <- confusionMatrix(tabNB)

#log loss scores
yNB    <- testNB$Position
p_trueNB <- postNB[cbind(1:nrow(testNB), match(as.character(yNB), colnames(postNB)))]
loglossNB <- -mean(log(pmax(p_trueNB)))
```


```{r,echo = FALSE}
#null model
pi_trainNB <- prop.table(table(trainNB$Position))
maj <- names(which.max(pi_trainNB))
loglossnullNB <- -mean(log(pi_trainNB[as.character(yNB)]))
pred_null <- factor(rep(maj, nrow(testNB)), levels = levels(yNB))
#null prediction matrix
tab_null <- table(predicted = pred_null, actual = yNB)
tab_null

```
The prediction matrix for the null model follows the earlier count and prior, showing that Catcher was the most populated position in the data set.

```{r,echo = FALSE}
# null confusion matrix
cm_nullNB <- confusionMatrix(table(factor(rep(maj, nrow(testNB)), levels = levels(yNB)), yNB))
```
The Confusion matrix for the null model.

```{r, echo = FALSE}
# prints metric scores for comparison between null and prediction
results <- data.frame(
  Model    = c("Naive Bayes", "No-skill"),
  Accuracy = c(cmNB$overall["Accuracy"], cm_nullNB$overall["Accuracy"]),
  BalAcc   = c(mean(cmNB$byClass[, "Balanced Accuracy"]),
               mean(cm_nullNB$byClass[, "Balanced Accuracy"])),       # baseline 0.5
  Kappa    = c(cmNB$overall["Kappa"], cm_nullNB$overall["Kappa"]),      # baseline 0
  LogLoss  = c(loglossNB, loglossnullNB)
)
results
```
## Comparison between Naive Bayes and Null model

# 5. Discussion and Real World Implications

# 6. Citations and Data Link
The data is linked [here](https://sabr.app.box.com/s/y1prhc795jk8zvmelfd3jq7tl389y6cd). We downloaded `Fielding.csv`, `batting.csv`, and `Teams.csv`.

# 7. Appendix

## Data cleaning
```r
pos     <- read.csv("Fielding.csv")
batting <- read.csv("batting.csv")
team    <- read.csv("Teams.csv")
# Getting highest position played for players
primary_pos <- pos %>%
  group_by(playerID, yearID, POS) %>%
  summarise(G = sum(G), .groups = "drop") %>%
  group_by(playerID, yearID) %>%
  slice_max(G, n = 1, with_ties = FALSE) %>%
  select(playerID, yearID, POS)

# Merge Position into Player and Year to condense data
stats <- merge(batting, primary_pos,
               by = c("playerID", "yearID"), all.x = TRUE)

# Keeps only American and National League
stats_clean <- stats[stats$lgID %in% c("AL", "NL"), ]
stats_clean <- stats_clean[!is.na(stats_clean$POS), ]


#####################################################
####### REGRESSION !!!!!!!!!
## Separates by team for regression prediction. 

team_clean <- team[team$lgID %in% c("AL","NL"),] # Keeps only AL and NL
team_clean$X1B <- team_clean$H - team_clean$X2B - team_clean$X3B - team_clean$HR # Creates X1B
team_clean$HBPxBB <- team_clean$BB + team_clean$HBP # Creates HBPxBB


# Takes only the necessary data
team_clean <- team_clean[,c("yearID", "teamID","G", "R", "X1B", "X2B", "X3B", 
                            "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")] 

#Divides stats by games to establish per game base
stats <- c("R", "X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF")
team_clean[stats] <- team_clean[stats] / team_clean$G

team_clean <- team_clean[!is.na(team_clean$SF), ]# Removes NaN values from SF
team_clean <- team_clean[!is.na(team_clean$HBPxBB), ]# Removes NaN values from HBPxBB

# List of predictors for training
num_vars <- c("X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB") 


###################################################

# CLASSIFICATION AND BAYES!!!!!!!!!!!!!
# Keeps only At Bats > 100
pos_data <- stats_clean[stats_clean$AB >= 100, ] 

# Changes The infield positions into a common infield tab
for(i in 1:length(pos_data$POS)){
  if(pos_data$POS[i] == "1B" || 
     pos_data$POS[i] == "2B" ||
     pos_data$POS[i] == "3B" ||
     pos_data$POS[i] == "SS" ){
    pos_data$POS[i] = "IF"
  }
}
# For easier referral
position_names <- c(
  "IF" = "Infield",
  "C"  = "Catcher",
  "OF" = "Outfield",
  "P"  = "Pitcher"
)

#Factors Positions into levels
pos_data$Position <- factor(position_names[pos_data$POS])

#Keep only necessary data
pos_data <- pos_data[, c("playerID", "yearID", "Position",
                         "H", "HR", "BB", "SO", "SB", "AB")]

# creates scaled pos_data for Classification comparison and naive Bayes
pos_scale <- pos_data

pos_scale$BA <- pos_data$H  / pos_data$AB
pos_scale$HRA <- pos_data$HR / pos_data$AB
pos_scale$BBA <- pos_data$BB / pos_data$AB
pos_scale$SOA <- pos_data$SO / pos_data$AB

# Factors Pos_Scaled
pos_scale$Position <- factor(position_names[pos_data$Position])

#Adds scaled variables into pos_scaled
pos_scale <- pos_scale[, c("playerID", "yearID", "Position",
                                  "BA", "HRA", "BBA", "SOA")]
 
```

## Scale vs Unscale Distance
```r
#creates table for unscaled distances
d2 <-(team_clean[1, num_vars] - team_clean[2, num_vars])^2
round(rbind(squared_diff = unlist(d2), percent = 100*unlist(d2) / sum(d2)),2)

# table for scaled distances
z <- scale(team_clean[, num_vars])
d2z <- (z[1,] - z[2,])^2
round(rbind(squared_diff = d2z, percent = 100*d2z/sum(d2z)),2)
```
## KNN Regression
```r

#splits training/testing 70/30
set.seed(4654)
index <- sample(1:nrow(team_clean), round(nrow(team_clean) * 0.7))
training_df <- scale(team_clean[index, num_vars])
testing_df  <- scale(team_clean[-index, num_vars],
                     center = attr(training_df, "scaled:center"),
                     scale  = attr(training_df, "scaled:scale"))

train_pred <- team_clean$R[index]
test_pred <- team_clean$R[-index]

# Cross Validation for best K
set.seed(4654)
folds <- sample(rep(1:10, length.out = nrow(training_df)))

cv_rmse <- function(k) {
  sq_err <- unlist(lapply(1:10, function(f) {
    tr_x <- training_df[folds != f, ]
    te_x <- training_df[folds == f, ]
    tr_y <- train_pred[folds != f]
    te_y <- train_pred[folds == f]
    pred <- knn.reg(train = tr_x, test = te_x, y = tr_y, k = k)$pred
    (te_y - pred)^2
  }))
  sqrt(mean(sq_err))
}

k_cv <- 1:80
cv_results <- data.frame(k = k_cv, RMSE = sapply(k_cv, cv_rmse))
k_best <- cv_results$k[which.min(cv_results$RMSE)]

#Training knn for values
train_rmse <- function(k) {
  pred <- knn.reg(train = training_df, test = training_df, y = train_pred, k = k)$pred
  sqrt(mean((train_pred - pred)^2))
}
cv_results$Train_RMSE <- sapply(cv_results$k, train_rmse)

# plot showing best k and training rmse results
ggplot(cv_results, aes(x = k)) +
  geom_line(aes(y = RMSE, color = "10-fold CV"), linewidth = 1) +
  geom_line(aes(y = Train_RMSE, color = "Training"), linewidth = 1) +
  geom_point(data = cv_results[cv_results$k == k_best, ],
             aes(y = RMSE), color = "red", size = 4) +
  scale_color_manual(values = c("10-fold CV" = "#8B2346", "Training" = "gray50")) +
  labs(title = paste0("Cross-validated choice:  k = ", k_best),
       x = "k", y = "RMSE", color = NULL) +
  theme_bw(base_size = 14)


top_k <- head(cv_results[order(cv_results$RMSE), ], 5)
top_k$RMSE   <- round(top_k$RMSE, 3)
rownames(top_k) <- NULL
top_k # plot showing top 5 ks and their RMSE


final_pred <- knn.reg(train = training_df, test = testing_df,
                      y = train_pred, k = k_best)$pred

rmse      <- function(a, b) sqrt(mean((a - b)^2))
mae       <- function(a, b) mean(abs(a - b))
r_squared <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)

null_prediction <- rep(mean(train_pred), length(test_pred))

results <- data.frame(
  Model = c("Null (training mean)", paste0("KNN (k = ", k_best, ")")),
  RMSE  = c(rmse(test_pred, null_prediction), rmse(test_pred, final_pred)),
  MAE   = c(mae(test_pred, null_prediction),  mae(test_pred, final_pred)),
  R2    = c(r_squared(test_pred, null_prediction), r_squared(test_pred, final_pred))
)
results[, -1] <- round(results[, -1], 3)
results
```
## KNN Classification

## Naive Bayes
```r
H <- table(pos_scale$Position)# Establish count
prior <- prop.table(H)# establish prior
rbind(Count = H, Prior = round(prior, 4))# binds count and prior to one table

# Splits train/test 70/30
set.seed(4654)
idx <- unlist(lapply(split(1:nrow(pos_scale), pos_scale$Position),
                     function(i) sample(i, round(0.7 * length(i)))))
trainNB <- pos_scale[idx,]
testNB <- pos_scale[-idx,]
# gives categorical likelihood tables
fitNB <- naiveBayes(Position ~ BA + HRA + BBA + SOA, data = trainNB) 
fitNB$tables 

# class prediction and its posterior
mapNB <- predict(fitNB, newdata = testNB, type = "class") 
postNB <- predict(fitNB, newdata = testNB, type = "raw")

#data frame of predictions and posteriors
pred_dfNB <- data.frame(
  predicted = mapNB,
  posterior   = round(apply(postNB, 1, max), 4)
)
head(pred_dfNB, 10)
# Prediction matrix
tabNB <- table(predicted = predict(fitNB, newdata = testNB), actual = testNB$Position)
tabNB

# Confusion matrix 
cmNB <- confusionMatrix(tabNB)

#log loss scores
yNB    <- testNB$Position
p_trueNB <- postNB[cbind(1:nrow(testNB), match(as.character(yNB), colnames(postNB)))]
loglossNB <- -mean(log(pmax(p_trueNB)))

#null model
pi_trainNB <- prop.table(table(trainNB$Position))
maj <- names(which.max(pi_trainNB))
loglossnullNB <- -mean(log(pi_trainNB[as.character(yNB)]))
pred_null <- factor(rep(maj, nrow(testNB)), levels = levels(yNB))
#null prediction matrix
tab_null <- table(predicted = pred_null, actual = yNB)
tab_null

# null confusion matrix
cm_nullNB <- confusionMatrix(table(factor(rep(maj, nrow(testNB)), levels = levels(yNB)), yNB))

# prints metric scores for comparison between null and prediction
results <- data.frame(
  Model    = c("Naive Bayes", "No-skill"),
  Accuracy = c(cmNB$overall["Accuracy"], cm_nullNB$overall["Accuracy"]),
  BalAcc   = c(mean(cmNB$byClass[, "Balanced Accuracy"]),
               mean(cm_nullNB$byClass[, "Balanced Accuracy"])),       # baseline 0.5
  Kappa    = c(cmNB$overall["Kappa"], cm_nullNB$overall["Kappa"]),      # baseline 0
  LogLoss  = c(loglossNB, loglossnullNB)
)
results

```

## Comparison