---
title: "CMDA-4654  \n Group Exercise 1 Part A"
subtitle: "Group 17"
author: "Jacob Carlson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    highlight: kate
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---

```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here
library(readr)
library(dplyr)
library(ggplot2)

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE, # Turn off warnings
                      R.options = list(width = 100))

```

\clearpage

# 1. Teammate Introduction

My teammate's name is Sophia. She's from Fall's Church City, VA. Her favorite food to eat in Blacksburg is Kuma Moon and her favorite hobby is watching TV shows with her friends.

# 2. Data Introduction

The area of data we chose to explore is sports. The data set is sourced from the Society for American Baseball Research's Lahman Baseball Database.

The data is linked [here](https://sabr.app.box.com/s/y1prhc795jk8zvmelfd3jq7tl389y6cd). We downloaded `FieldingPost.csv` and `batting.csv`.

summarization of what the data corresponds to

## Data Dictionary

`FieldingPost.csv` contains post-season fielding data with the following variables:

| Variable   | Description                                           |
|------------|-------------------------------------------------------|
| `playerID` | Player ID code                                        |
| `yearID`   | Year                                                  |
| `stint`    | Player's stint (order of appearances within a season) |
| `teamID`   | Team                                                  |
| `lgID`     | League                                                |
| `G`        | Games                                                 |
| `AB`       | At Bats                                               |
| `R`        | Runs                                                  |
| `H`        | Hits                                                  |
| `2B`       | Doubles                                               |
| `3B`       | Triples                                               |
| `HR`       | Home Runs                                             |
| `RBI`      | Runs Batted In                                        |
| `SB`       | Stolen Bases                                          |
| `CS`       | Caught Stealing                                       |
| `BB`       | Base on Balls                                         |
| `SO`       | Strikeouts                                            |
| `IBB`      | Intentional Walks                                     |
| `HBP`      | Hit by Pitch                                          |
| `SH`       | Sacrifice Hits                                        |
| `SF`       | Sacrifice Flies                                       |
| `GIDP`     | Grounded Into Double Plays                            |

`batting.csv` contains batting statistics with the following variables:

| Variable   | Description                                           |
|------------|-------------------------------------------------------|
| `playerID` | Player ID code                                        |
| `yearID`   | Year                                                  |
| `stint`    | Player's stint (order of appearances within a season) |
| `teamID`   | Team                                                  |
| `lgID`     | League                                                |
| `G`        | Games                                                 |
| `AB`       | At Bats                                               |
| `R`        | Runs                                                  |
| `H`        | Hits                                                  |
| `2B`       | Doubles                                               |
| `3B`       | Triples                                               |
| `HR`       | Home Runs                                             |
| `RBI`      | Runs Batted In                                        |
| `SB`       | Stolen Bases                                          |
| `CS`       | Caught Stealing                                       |
| `BB`       | Base on Balls                                         |
| `SO`       | Strikeouts                                            |
| `IBB`      | Intentional Walks                                     |
| `HBP`      | Hit by Pitch                                          |
| `SH`       | Sacrifice Hits                                        |
| `SF`       | Sacrifice Flies                                       |
| `GIDP`     | Grounded Into Double Plays                            |

# 3. Analysis and Discussion

```{r, include=FALSE}
# data cleaning code, not included in knitted file (included in appendix)

## Building data frames 

pos <- read.csv("FieldingPost.csv")
batting <- read.csv("batting.csv")

stats <- merge(batting, pos[, c("playerID", "POS")],
               by = "playerID", all.x = TRUE)# Add positions

stats_clean <- stats[stats$lgID == "AL" |
                    stats$lgID == "NL", ] # Focuses on American and National League

stats_clean <- stats_clean[!is.na(stats_clean$POS), ]  # Removes batting data without positions

stats_clean$BA <- ifelse(stats_clean$AB > 0, stats_clean$H /
                           stats_clean$AB, NA) ## Adds batting average

stats_wo_pitch <- stats_clean[stats_clean$POS != "P", ] # Removes pitchers for different rules

# league and year 
by_league_W_pitch <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ lgID + yearID,
                                data = stats_clean, FUN = sum) 
# Adds Batting Average
by_league_W_pitch$BA <- by_league_W_pitch$H / by_league_W_pitch$AB

# league and year no pitcher
by_league_Wo_pitch <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ lgID + yearID,
                                 data = stats_wo_pitch, FUN = sum)
# Adds Batting Average
by_league_Wo_pitch$BA <- by_league_Wo_pitch$H / by_league_Wo_pitch$AB # Adds Batting Average

# team and year
by_team <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ teamID + yearID,
                      data = stats_clean, FUN = sum) 
# Adds Batting Average
by_team$BA <- by_team$H / by_team$AB # Adds Batting Average

by_year <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ yearID,
                      data = stats_clean, FUN = sum) #year
# Adds Batting Average
by_year$BA <- by_year$H / by_year$AB # Adds Batting Average

# Position
by_pos <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ POS,
                     data = stats_clean, FUN = sum) 
# Adds Batting Average
by_pos$BA <- by_pos$H / by_pos$AB 

# rename positions 
position_names <- c(
  "1B" = "First Base",
  "2B" = "Second Base",
  "3B" = "Third Base",
  "C"  = "Catcher",
  "CF" = "Center Field",
  "LF" = "Left Field",
  "P"  = "Pitcher",
  "RF" = "Right Field",
  "SS" = "Shortstop"
)

by_pos$POS <- position_names[by_pos$POS]

#Position and Team
by_team_pos <- aggregate(cbind(AB, H, HR, BB, SO, BA) ~ teamID + POS, data = stats_clean, FUN = sum)  
# Adds Batting Average
by_team_pos$BA <- by_team_pos$H / by_team_pos$AB



by_team_pos$Position <- position_names[by_team_pos$POS]
```

## Tables

```{r, echo=FALSE}
# columns of interest
stat_cols <- c("AB", "R", "H", "HR", "SB", "BB", "SO", "BA")

# generates summary statistics
compact_summary <- function(df, cols) {
  
  variable_names <- c(
    AB = "At Bats",
    R = "Runs",
    H = "Hits",
    HR = "Home Runs",
    SB = "Stolen Bases",
    BB = "Base on Balls",
    SO = "Strikeouts",
    BA = "Batting Average"
  )
  
  data.frame(
    Variable = unname(variable_names[cols]),
    Minimum = sapply(df[cols], min, na.rm = TRUE),
    Mean = sapply(df[cols], mean, na.rm = TRUE),
    Maximum = sapply(df[cols], max, na.rm = TRUE),
    `Standard Deviation` = sapply(df[cols], sd, na.rm = TRUE),
    row.names = NULL
  )
}

# formats numbers in table
format_table <- function(table_name) {
  table_name[, 2:5] <- lapply(
    table_name[, 2:5],
    function(x) format(round(x, 3), big.mark = ",", trim = TRUE)
  )
  
  return(table_name)
}

league_summary_W_pitch <- compact_summary(
  by_league_W_pitch, 
  stat_cols
)

knitr::kable(
  format_table(league_summary_W_pitch),
  caption = "Statistics Aggregated by League (With Pitchers)"
)
```
discussion

```{r, echo=FALSE}
league_summary_Wo_pitch <- compact_summary(
  by_league_Wo_pitch, 
  stat_cols
)

knitr::kable(
  format_table(league_summary_Wo_pitch),
  caption = "Statistics Aggregated by League (Without Pitchers)"
)

```
discussion

## Plots

```{r, echo=FALSE}

ggplot(by_pos, aes(x = POS, y = BA)) +
  geom_bar(stat = "identity", fill = "maroon") +
  labs(
    title = "Batting Average by Position",
    x = "Position",
    y = "Batting Average"
  )  +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
discussion

```{r, echo=FALSE}
ggplot(by_team_pos, aes(x = Position, y = SO)) +
  geom_boxplot(fill="maroon") +
  labs(
    title = "Strikeouts Across Teams by Position",
    x = "Position",
    y = "Strikeouts"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
discussion

```{r, echo=FALSE}
ggplot(by_team, aes(x=H, y=HR)) +
  geom_point(col="maroon") +
  labs(
    title="Hits vs Home Runs Across Teams",
    x = "Hits",
    y = "Home Runs"
  )
```
discussion

```{r, echo=FALSE}
ggplot(by_team, aes(x=SO, y=BA)) +
  geom_point(col="maroon") +
  labs(
    title="Strikeouts vs Batting Average Across Teams",
    x = "Strikeouts",
    y = "Batting Averages"
  )

```

# 4. Citations

Lahman Baseball Database – Society for American Baseball Research. (2025). Retrieved September 12, 2026, from <https://sabr.org/lahman-database/>

# 5. Appendix

## Data Cleaning Code

```r
## Building data frames 

pos <- read.csv("FieldingPost.csv")
batting <- read.csv("batting.csv")

stats <- merge(batting, pos[, c("playerID", "POS")],
               by = "playerID", all.x = TRUE)# Add positions

stats_clean <- stats[stats$lgID == "AL" |
                    stats$lgID == "NL", ] # Focuses on American and National League

stats_clean <- stats_clean[!is.na(stats_clean$POS), ]  # Removes batting data without positions

stats_clean$BA <- ifelse(stats_clean$AB > 0, stats_clean$H /
                           stats_clean$AB, NA) ## Adds batting average

stats_wo_pitch <- stats_clean[stats_clean$POS != "P", ] # Removes pitchers for different rules

# league and year 
by_league_W_pitch <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ lgID + yearID,
                                data = stats_clean, FUN = sum) 
# Adds Batting Average
by_league_W_pitch$BA <- by_league_W_pitch$H / by_league_W_pitch$AB

# league and year no pitcher
by_league_Wo_pitch <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ lgID + yearID,
                                 data = stats_wo_pitch, FUN = sum)
# Adds Batting Average
by_league_Wo_pitch$BA <- by_league_Wo_pitch$H / by_league_Wo_pitch$AB # Adds Batting Average

# team and year
by_team <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ teamID + yearID,
                      data = stats_clean, FUN = sum) 
# Adds Batting Average
by_team$BA <- by_team$H / by_team$AB # Adds Batting Average

by_year <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ yearID,
                      data = stats_clean, FUN = sum) #year
# Adds Batting Average
by_year$BA <- by_year$H / by_year$AB # Adds Batting Average

# Position
by_pos <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ POS,
                     data = stats_clean, FUN = sum) 
# Adds Batting Average
by_pos$BA <- by_pos$H / by_pos$AB 

# rename positions 
position_names <- c(
  "1B" = "First Base",
  "2B" = "Second Base",
  "3B" = "Third Base",
  "C"  = "Catcher",
  "CF" = "Center Field",
  "LF" = "Left Field",
  "P"  = "Pitcher",
  "RF" = "Right Field",
  "SS" = "Shortstop"
)

by_pos$POS <- position_names[by_pos$POS]

#Position and Team
by_team_pos <- aggregate(cbind(AB, H, HR, BB, SO, BA) ~ teamID + POS, data = stats_clean, FUN = sum)  
# Adds Batting Average
by_team_pos$BA <- by_team_pos$H / by_team_pos$AB


by_team_pos$Position <- position_names[by_team_pos$POS]
```

## Code for Tables

```r
# columns of interest
stat_cols <- c("AB", "R", "H", "HR", "SB", "BB", "SO", "BA")

# generates summary statistics
compact_summary <- function(df, cols) {
  
  variable_names <- c(
    AB = "At Bats",
    R = "Runs",
    H = "Hits",
    HR = "Home Runs",
    SB = "Stolen Bases",
    BB = "Base on Balls",
    SO = "Strikeouts",
    BA = "Batting Average"
  )
  
  data.frame(
    Variable = unname(variable_names[cols]),
    Minimum = sapply(df[cols], min, na.rm = TRUE),
    Mean = sapply(df[cols], mean, na.rm = TRUE),
    Maximum = sapply(df[cols], max, na.rm = TRUE),
    `Standard Deviation` = sapply(df[cols], sd, na.rm = TRUE),
    row.names = NULL
  )
}

# formats numbers in table
format_table <- function(table_name) {
  table_name[, 2:5] <- lapply(
    table_name[, 2:5],
    function(x) format(round(x, 3), big.mark = ",", trim = TRUE)
  )
  
  return(table_name)
}

league_summary_W_pitch <- compact_summary(
  by_league_W_pitch, 
  stat_cols
)

knitr::kable(
  format_table(league_summary_W_pitch),
  caption = "Statistics Aggregated by League (With Pitchers)"
)

league_summary_Wo_pitch <- compact_summary(
  by_league_Wo_pitch, 
  stat_cols
)

knitr::kable(
  format_table(league_summary_Wo_pitch),
  caption = "Statistics Aggregated by League (Without Pitchers)"
)

```
## Code for Plots

```r

ggplot(by_pos, aes(x = POS, y = BA)) +
  geom_bar(stat = "identity", fill = "maroon") +
  labs(
    title = "Batting Average by Position",
    x = "Position",
    y = "Batting Average"
  )  +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplot(by_team_pos, aes(x = Position, y = SO)) +
  geom_boxplot(fill="maroon") +
  labs(
    title = "Strikeouts Across Teams by Position",
    x = "Position",
    y = "Strikeouts"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggplot(by_team, aes(x=H, y=HR)) +
  geom_point(col="maroon") +
  labs(
    title="Hits vs Home Runs Across Teams",
    x = "Hits",
    y = "Home Runs"
  )

ggplot(by_team, aes(x=SO, y=BA)) +
  geom_point(col="maroon") +
  labs(
    title="Strikeouts vs Batting Average Across Teams",
    x = "Strikeouts",
    y = "Batting Averages"
  )
```