---
title: "CMDA-4654  \n Group Exercise 1B"
author: "Sophia Gavasheli and Jacob Carlson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    highlight: kate
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---
  

```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here
library(readr)
library(dplyr)
library(ggplot2)
library(e1071)
library(caret)

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE, # Turn off warnings
                      R.options = list(width = 100))

library(here)
setwd(here()) # sets path to root of RProject

```

\clearpage



# 1. Reintroduction
Can we predict a team's seasonal runs based on it's offensive stats(KNN Regression)

Can we predict a players position based on their offensive stats(KNN Classification and Naive Bayes)
# Executive Summary and Research Questions 
For our KNN regression analyis, we wanted to predict a teams seasonal runs based off of its offensive stats. We will use 
```{r,echo = FALSE}
pos     <- read.csv("Fielding.csv")
batting <- read.csv("batting.csv")
team    <- read.csv("Teams.csv")
# Getting highest position played for players
primary_pos <- pos %>%
  group_by(playerID, yearID, POS) %>%
  summarise(G = sum(G), .groups = "drop") %>%
  group_by(playerID, yearID) %>%
  slice_max(G, n = 1, with_ties = FALSE) %>%
  select(playerID, yearID, POS)

# Merge Position into Player and Year to condense data
stats <- merge(batting, primary_pos,
               by = c("playerID", "yearID"), all.x = TRUE)

# Keeps only American and National League
stats_clean <- stats[stats$lgID %in% c("AL", "NL"), ]
stats_clean <- stats_clean[!is.na(stats_clean$POS), ]


#####################################################
####### REGRESSION !!!!!!!!!
## Seperates by team for regression prediction. 

team_clean <- team[team$lgID %in% c("AL","NL"),]
team_clean$X1B <- team_clean$H - team_clean$X2B - team_clean$X3B - team_clean$HR
team_clean$HBPxBB <- team_clean$BB + team_clean$HBP
team_clean <- team_clean[,c("yearID", "teamID", "R", "X1B", "X2B", "X3B", 
                            "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")]
team_clean <- team_clean[!is.na(team_clean$SF), ]
team_clean <- team_clean[!is.na(team_clean$HBPxBB), ]
# Creates second frame to scale data per at-bat
team_scaled <- team_clean

###################################################

# CLASSIFICATION AND BAYES!!!!!!!!!!!!!
# Uses 100 minimum at bats to remove extreme cases
pos_data <- stats_clean[stats_clean$AB >= 100, ]

# Changes The infield positions into a common infield tab
for(i in 1:length(pos_data$POS)){
  if(pos_data$POS[i] == "1B" || 
     pos_data$POS[i] == "2B" ||
     pos_data$POS[i] == "3B" ||
     pos_data$POS[i] == "SS" ){
    pos_data$POS[i] = "IF"
  }
}
position_names <- c(
  "IF" = "Infield",
  "C"  = "Catcher",
  "OF" = "Outfield",
  "P"  = "Pitcher"
)
pos_data$Position <- factor(position_names[pos_data$POS])


pos_data <- pos_data[, c("playerID", "yearID", "Position",
                         "H", "HR", "BB", "SO", "SB", "AB")]
NB <- pos_data

# creates scaled pos_data for Classification comparison and naive Bayes
pos_scale <- pos_data

pos_scale$BA <- pos_data$H  / pos_data$AB
pos_scale$HRA <- pos_data$HR / pos_data$AB
pos_scale$BBA <- pos_data$BB / pos_data$AB
pos_scale$SOA <- pos_data$SO / pos_data$AB

pos_scale$Position <- factor(position_names[pos_data$Position])

pos_scale <- pos_scale[, c("playerID", "yearID", "Position",
                                  "BA", "HRA", "BBA", "SOA")]
```




## Naive Bayes
```{r,echo = FALSE}
H <- table(pos_scale$Position)# Establish count
prior <- prop.table(H)# establish prior
rbind(Count = H, Prior = round(prior, 4))# binds count and prior to one table

```
A basic Count/Prior table has the catcher and outfield positions being the most populated.
```{r,echo = FALSE}
set.seed(4654)
idx <- unlist(lapply(split(1:nrow(pos_scale), pos_scale$Position),
                     function(i) sample(i, round(0.75 * length(i)))))
trainNB <- pos_scale[idx,]
testNB <- pos_scale[-idx,]
fitNB <- naiveBayes(Position ~ BA + HRA + BBA + SOA, data = trainNB)
fitNB$tables
```
All of the predictors are numerical, and therefor modeled with  Gaussian class-conditional densities. Furthermore, there is no need to consider Laplace smoothing.
```{r,echo = FALSE}


mapNB <- predict(fitNB, newdata = testNB, type = "class")
postNB <- predict(fitNB, newdata = testNB, type = "raw")

pred_dfNB <- data.frame(
  predicted = mapNB,
  posterior   = round(apply(postNB, 1, max), 4)
)
head(pred_dfNB, 10)
tabNB <- table(predicted = predict(fitNB, newdata = testNB), actual = testNB$Position)
tabNB
```
I have produced the first 10 hard predictions and soft posteriors. Below that is the accompanying prediction matrix.

```{r,echo = FALSE}
cmNB <- confusionMatrix(tabNB)
yNB    <- testNB$Position
p_trueNB <- postNB[cbind(1:nrow(testNB), match(as.character(yNB), colnames(postNB)))]
loglossNB <- -mean(log(pmax(p_trueNB)))
```


```{r,echo = FALSE}
pi_trainNB <- prop.table(table(trainNB$Position))
maj <- names(which.max(pi_trainNB))
loglossnullNB <- -mean(log(pi_trainNB[as.character(yNB)]))
pred_null <- factor(rep(maj, nrow(testNB)), levels = levels(yNB))
tab_null <- table(predicted = pred_null, actual = yNB)
tab_null

```
The prediction matrix for the null model follows the earlier count and prior, showing that Catcher was the most populated position in the data set.

```{r,echo = FALSE}

cm_nullNB <- confusionMatrix(table(factor(rep(maj, nrow(testNB)), levels = levels(yNB)), yNB))
```
The Confusion matrix for the null model.

```{r, echo = FALSE}
######################################
# Naive Bayes

########################


results <- data.frame(
  Model    = c("Naive Bayes", "No-skill"),
  Accuracy = c(cmNB$overall["Accuracy"], cm_nullNB$overall["Accuracy"]),
  BalAcc   = c(mean(cmNB$byClass[, "Balanced Accuracy"]),
               mean(cm_nullNB$byClass[, "Balanced Accuracy"])),       # baseline 0.5
  Kappa    = c(cmNB$overall["Kappa"], cm_nullNB$overall["Kappa"]),      # baseline 0
  LogLoss  = c(loglossNB, loglossnullNB)
)
results
```
Comparison between Naive Bayes and Null model


