---
title: "CMDA-4654  \n Group Exercise 1B"
author: "Sophia Gavasheli and Jacob Carlson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    highlight: kate
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---
  

```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here
library(readr)
library(dplyr)
library(ggplot2)
library(e1071)
library(caret)
library(FNN)

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE, # Turn off warnings
                      R.options = list(width = 100))

library(here)
setwd(here()) # sets path to root of RProject

```

\clearpage



# 1. Reintroduction
We are Sophia Gavasheli and Jacob Carlson, and we our group 17. Our chosen domain is the sports category.

# 2. Executive Summary and Research Questions 

## KNN Regression
We are using a KNN regression model to predict the runs a team scored in any given game. Runs will be predicted by the following variables.

| Variable   | Description                                           |
|------------|-------------------------------------------------------|
| `AB`       | At Bats                                               |
| `1b`       | Singles                                               |
| `2B`       | Doubles                                               |
| `3B`       | Triples                                               |
| `HR`       | Home Runs                                             |
| `SB`       | Stolen Bases                                          |
| `CS`       | Caught Stealing                                       |
| `BB`       | Base on Balls                                         |
| `SO`       | Strikeouts                                            |
| `HBP`      | Hit by Pitch                                          |
| `SH`       | Sacrifice Hits                                        |
| `SF`       | Sacrifice Flies                                       |

The data frame that these variables were taken from was not used in part A of our exercise, but is still part of the larger data set we pulled from. Using runs as a metric was an explored idea in part A, but was not used in our final report, so we decided to further explore it here, since it looked like it could have a relationship with the other variables like home runs and at bats.

## KNN Classification and Naive Bayes
For our KNN Classification and Naive Bayes models, we chose to predict a player's position using the following offensive stats:

| Variable   | Description                                           |
|------------|-------------------------------------------------------|
| `BA`       | Batting Average                                       |
| `HRA`      | Home Runs Average                                     |
| `BBA`      | Base on Balls Average                                 |
| `SOA`      | Strikeout Average                                     |

We standardized this data to ensure KNN and Naive Bayes were using the same inputs. We actually spent a lot of our part A analysis on position metric, so exploring further with the two models was a natural part B. 

In part A, we used the data frames `fieldingpost.csv` and `batting.csv`. Fielding-post only included post season stats(playoffs), and wasn't big enough for our models, so we switched to `fielding.csv`, which contains the entire seasons. We also used `Teams.csv` for our KNN regression, as it contained all of the data used in our model with minimal cleaning and merging between data sets.


```{r,echo = FALSE}
pos     <- read.csv("Fielding.csv")
batting <- read.csv("batting.csv")
team    <- read.csv("Teams.csv")
# Getting highest position played for players
primary_pos <- pos %>%
  group_by(playerID, yearID, POS) %>%
  summarise(G = sum(G), .groups = "drop") %>%
  group_by(playerID, yearID) %>%
  slice_max(G, n = 1, with_ties = FALSE) %>%
  select(playerID, yearID, POS)

# Merge Position into Player and Year to condense data
stats <- merge(batting, primary_pos,
               by = c("playerID", "yearID"), all.x = TRUE)

# Keeps only American and National League
stats_clean <- stats[stats$lgID %in% c("AL", "NL"), ]
stats_clean <- stats_clean[!is.na(stats_clean$POS), ]


####### REGRESSION 
## Separates by team for regression prediction. 

team_clean <- team[team$lgID %in% c("AL","NL"),] # Keeps only AL and NL
team_clean$X1B <- team_clean$H - team_clean$X2B - team_clean$X3B - team_clean$HR # Creates X1B
team_clean$HBPxBB <- team_clean$BB + team_clean$HBP # Creates HBPxBB


# Takes only the necessary data
team_clean <- team_clean[,c("yearID", "teamID","G", "R", "X1B", "X2B", "X3B", 
                            "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")] 

#Divides stats by games to establish per game base
stats <- c("R", "X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")
team_clean[stats] <- team_clean[stats] / team_clean$G

team_clean <- team_clean[!is.na(team_clean$SF), ]# Removes NaN values from SF
team_clean <- team_clean[!is.na(team_clean$HBPxBB), ]# Removes NaN values from HBPxBB

# List of predictors for training
num_vars <- c("X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB") 


##### CLASSIFICATION AND BAYES
# Keeps only At Bats > 100
pos_data <- stats_clean[stats_clean$AB >= 100, ] 

# Changes The infield positions into a common infield tab
for(i in 1:length(pos_data$POS)){
  if(pos_data$POS[i] == "1B" || 
     pos_data$POS[i] == "2B" ||
     pos_data$POS[i] == "3B" ||
     pos_data$POS[i] == "SS" ){
    pos_data$POS[i] = "IF"
  }
}
# For easier referral
position_names <- c(
  "IF" = "Infield",
  "C"  = "Catcher",
  "OF" = "Outfield",
  "P"  = "Pitcher"
)

#Factors Positions into levels
pos_data$Position <- factor(position_names[pos_data$POS])

#Keep only necessary data
pos_data <- pos_data[, c("playerID", "yearID", "Position",
                         "H", "HR", "BB", "SO", "SB", "AB")]


# creates scaled pos_data for Classification comparison and naive Bayes
pos_scale <- pos_data

pos_scale$BA <- pos_data$H  / pos_data$AB
pos_scale$HRA <- pos_data$HR / pos_data$AB
pos_scale$BBA <- pos_data$BB / pos_data$AB
pos_scale$SOA <- pos_data$SO / pos_data$AB

#Adds scaled variables into pos_scaled
pos_scale <- pos_scale[, c("playerID", "yearID", "Position",
                                  "BA", "HRA", "BBA", "SOA")]
# List of Predictors for pos_scaled
feats <- c("BA", "HRA", "BBA", "SOA") 
```
\clearpage

# 3. Methodology & Modeling Setup

## Data Preprocessing

All of our predictors were numerical, so there was no need to handle categorical leveling or do any smoothing to our data.

We did a 70/30 training-test split since the data sets were very large, and we preferred to keep the majority in our training.

### KNN regression

For our KNN regression, `Teams.csv` includes a great many more predictors than what was used.
We built the data frame `team_clean`, which keeps only the American and National leagues, removes `NaN` rows, and takes in only our predictors, `TeamId`, `yearID`, and our prediction variable `Runs.` We also combined the `HBP` and `BB` predictors into one, as they provide the same outcome in a game. We also divided all of our predictors by games to establish the per game basis for our prediction.

We standardized our data set under the basic z-score scaling. As mentioned above, the predictor `HBPxBB` was a sum of two predictors, and was unfairly weighted against our other data. Furthermore, given the nature of the game, some of our predictors had a reported 0 weight on our distance, so standardizing was necessary.

The table below shows why scaling matters: for two team-seasons, the unscaled distance is dominated by a few large-valued predictors, while after z-scoring each predictor contributes a more balanced share.

```{r, echo = FALSE}
# table format
tbl <- function(x, cap, digits = 3, rn = FALSE, ...) {
  knitr::kable(x, booktabs = TRUE, caption = cap, digits = digits,
               position = "H", row.names = rn, ...)
}

d2  <- unlist((team_clean[1, num_vars] - team_clean[2, num_vars])^2)
z   <- scale(team_clean[, num_vars])
d2z <- (z[1, ] - z[2, ])^2
demo <- rbind("Raw (share of squared distance, pct)"          = 100 * d2  / sum(d2),
              "Z-scored (share of squared distance, pct)"     = 100 * d2z / sum(d2z))
tbl(demo, "Share of squared Euclidean distance contributed by each predictor (first two games).",
    digits = 1, rn = TRUE)
```

### KNN and Naive Bayes Classification

For our KNN classification and Naive Bayes models, we first created `primary_pos`, which read through the player's data, found their most played position, and changed their other reported positions to the highest one. We then combined `primary_pos`with batting, based on their `playerId` and the `yearID`, so that we could compare the positions to their offensive stats. We also took only the American and National league players, following our analysis in part A. 

We wanted to limit classification responses, so we combined the infield positions (1B,2B,3B, SS) into an `Infield` category. 

In baseball, different positions get different amounts of playtime. Furthermore, because we combined four possible positions into one, our data was even more skewed, and in the case of KNN classification, our distances would be poorly weighted. We therefore decided to scale all of our predictors based on At bats, providing the predictors described above(BA, HRA, BBA, SOA). While the standardizing would have less affect on our Naive Bayes model, since our goal is to compare it with our classification, we chose to scale the predictors before running either model.

```{r, echo = FALSE}
# table format
tbl <- function(x, cap, digits = 3, rn = FALSE, ...) {
  knitr::kable(x, booktabs = TRUE, caption = cap, digits = digits,
               position = "H", row.names = rn, ...)
}

d2  <- unlist((pos_scale[1, feats] - pos_scale[2, feats])^2)
z   <- scale(pos_scale[, feats])
d2z <- (z[1, ] - z[2, ])^2
demo <- rbind("Raw (share of squared distance, pct)"          = 100 * d2  / sum(d2),
              "Z-scored (share of squared distance, pct)"     = 100 * d2z / sum(d2z))
tbl(demo, "Share of squared Euclidean distance contributed by each predictor (first two team-seasons).",
    digits = 1, rn = TRUE)
```
\clearpage

# 4. Analysis, Tuning & Results
```{r,echo = FALSE}

#regression split 70/30
set.seed(4654)
idx_reg   <- sample(nrow(team_clean), round(0.7 * nrow(team_clean)))
train_raw <- team_clean[idx_reg, num_vars]
test_raw  <- team_clean[-idx_reg, num_vars]
train_y   <- team_clean$R[idx_reg]
test_y    <- team_clean$R[-idx_reg]

mu_reg      <- colMeans(train_raw)
sd_reg      <- apply(train_raw, 2, sd)
training_df <- scale(train_raw, center = mu_reg, scale = sd_reg)
testing_df  <- scale(test_raw,  center = mu_reg, scale = sd_reg)

# classification split (knn and nb)

set.seed(4654)
idx_cls <- unlist(lapply(split(seq_len(nrow(pos_scale)), pos_scale$Position),
                         function(i) sample(i, round(0.7 * length(i)))))
cls_train <- pos_scale[idx_cls, ]
cls_test  <- pos_scale[-idx_cls, ]

mu_cls  <- colMeans(cls_train[, feats])
sd_cls  <- apply(cls_train[, feats], 2, sd)
train_x <- scale(cls_train[, feats], center = mu_cls, scale = sd_cls)
test_x  <- scale(cls_test[, feats],  center = mu_cls, scale = sd_cls)

lv   <- levels(pos_scale$Position)
K    <- length(lv)
y_tr <- as.integer(cls_train$Position)
y_te <- as.integer(cls_test$Position)

# 
pi_train     <- prop.table(table(cls_train$Position))[lv]
null_entropy <- -sum(pi_train * log(pi_train))
```

##  KNN Regression

###  K Tuning with Cross Validation

We searched $k = 1, \ldots, 80$ using 10-fold CV on the training data only.

```{r fig.cap = "kNN regression tuning: cross-validated and training RMSE versus k. The red point marks the CV-optimal k. ",echo = FALSE}
set.seed(4654)
folds <- sample(rep(1:10, length.out = nrow(train_raw)))

# Scale inside every fold using ONLY that fold's fitting data (no leakage)
fold_data <- lapply(1:10, function(f) {
  tr <- train_raw[folds != f, ]
  va <- train_raw[folds == f, ]
  m  <- colMeans(tr)
  s  <- apply(tr, 2, sd)
  list(tr_x = scale(tr, center = m, scale = s),
       va_x = scale(va, center = m, scale = s),
       tr_y = train_y[folds != f],
       va_y = train_y[folds == f])
})

cv_rmse <- function(k) {test_x
  sq_err <- unlist(lapply(fold_data, function(d) {
    pred <- knn.reg(train = d$tr_x, test = d$va_x, y = d$tr_y, k = k)$pred
    (d$va_y - pred)^2
  }))
  sqrt(mean(sq_err))
}

train_rmse <- function(k) {
  pred <- knn.reg(train = training_df, test = training_df, y = train_y, k = k)$pred
  sqrt(mean((train_y - pred)^2))
}

k_cv       <- 1:80
cv_results <- data.frame(k = k_cv,
                         RMSE = sapply(k_cv, cv_rmse),
                         Train_RMSE = sapply(k_cv, train_rmse))
k_best <- cv_results$k[which.min(cv_results$RMSE)]

ggplot(cv_results, aes(x = k)) +
  geom_line(aes(y = RMSE, color = "10-fold CV"), linewidth = 1) +
  geom_line(aes(y = Train_RMSE, color = "Training"), linewidth = 1) +
  geom_point(data = cv_results[cv_results$k == k_best, ],
             aes(y = RMSE), color = "red", size = 4) +
  scale_color_manual(values = c("10-fold CV" = "#8B2346", "Training" = "orange")) +
  labs(
       x = "k", y = "RMSE of Runs per game", color = NULL) +
  theme_bw(base_size = 12)
```

The CV-optimal value is $k^* = `r k_best`$. Very small $k$ gives a near-zero training RMSE (each point is nearly its own neighbor) but a worse CV RMSE, which is overfitting. As $k$ becomes large the two curves rise together, which is underfitting.

```{r,echo = FALSE}
top_k <- head(cv_results[order(cv_results$RMSE), c("k", "RMSE", "Train_RMSE")], 5)
names(top_k) <- c("k", "CV RMSE", "Training RMSE")
tbl(top_k, paste0("Top five candidate values of k by cross-validated RMSE. The chosen optimum is k = ",
                  k_best, "."))
```


### Test-set performance versus the null model

```{r,echo=FALSE}
final_pred <- knn.reg(train = training_df, test = testing_df, y = train_y, k = k_best)$pred

rmse      <- function(a, b) sqrt(mean((a - b)^2))
mae       <- function(a, b) mean(abs(a - b))
r_squared <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)

null_prediction <- rep(mean(train_y), length(test_y))

results <- data.frame(
  Model = c("Null (training mean)", paste0("KNN (k = ", k_best, ")")),
  RMSE  = c(rmse(test_y, null_prediction), rmse(test_y, final_pred)),
  MAE   = c(mae(test_y, null_prediction),  mae(test_y, final_pred)),
  R2    = c(r_squared(test_y, null_prediction), r_squared(test_y, final_pred))
)

results[, -1] <- round(results[, -1], 3)

names(results)[4] <- "R-squared"
tbl(results, "kNN regression on the held-out test set versus the null (mean-only) model.")
```
**RMSE:** Our null model was off by about half of a run every game when predicting the training average. Our k = 9 was off by about a quarter of a run every game, which is about a 50% reduction compared to our null model.

**MAE:** Our null models' absolute average miss is a little under half a run every game. The k = 9 model was off a little under half that, also being a little more than a 50% reduction compared to the null.

**R Squared:** As expected, our null model takes 0% of the predictors into account for its predictions. However, our k = 9 model takes 82%, signifying that the 10 predictors are a strong representation of the runs scored.

\clearpage

##  KNN Classification
```{r, echo = FALSE}
# helpers
onehot  <- function(y_int) diag(K)[y_int, , drop = FALSE]
logloss <- function(P, y_int) -mean(log(pmax(P[cbind(seq_along(y_int), y_int)], 1e-15)))
brier   <- function(P, y_int) mean(rowSums((P - onehot(y_int))^2))
pred_from_P <- function(P) max.col(P, ties.method = "first")   # argmax rule

conf_tab <- function(pred_int, y_int)
  table(Predicted = factor(pred_int, levels = 1:K, labels = lv),
        Actual    = factor(y_int,    levels = 1:K, labels = lv))

# One-versus-all metrics per class from a confusion matrix
per_class <- function(cm) {
  TP <- diag(cm); FP <- rowSums(cm) - TP; FN <- colSums(cm) - TP
  TN <- sum(cm) - TP - FP - FN
  sens <- TP / (TP + FN); spec <- TN / (TN + FP)
  data.frame(Class = colnames(cm),
             Sensitivity = sens,
             Specificity = spec,
             Precision   = ifelse(TP + FP == 0, NA, TP / (TP + FP)),
             F1          = 2 * TP / (2 * TP + FP + FN),
             BalAcc      = (sens + spec) / 2)
}

# Overall summary
summ <- function(P, pred_int, y_int) {
  cm <- conf_tab(pred_int, y_int); pc <- per_class(cm); n <- sum(cm)
  po <- sum(diag(cm)) / n
  pe <- sum(rowSums(cm) * colSums(cm)) / n^2
  c(Accuracy = po,
    BalAcc   = mean(pc$BalAcc),
    MacroF1  = mean(pc$F1),
    Kappa    = (po - pe) / (1 - pe),
    LogLoss  = logloss(P, y_int),
    Brier    = brier(P, y_int))
}
```

```{r,echo = FALSE}
k_grid <- seq(1, 51, by = 2)
kmax   <- max(k_grid)


nbr_labels <- function(train_mat, query_mat, y_int, kmax) {
  nn <- FNN::knnx.index(train_mat, query_mat, k = kmax)
  matrix(y_int[nn], nrow = nrow(nn))
}

# Smoothed vote fractions: (votes + 1) / (k + K)
prob_k <- function(NL, k) {
  votes <- sapply(seq_len(K), function(cl) rowSums(NL[, seq_len(k), drop = FALSE] == cl))
  P <- (votes + 1) / (k + K)
  colnames(P) <- lv
  P
}

#10-fold CV
set.seed(4654)
folds_c <- sample(rep(1:10, length.out = nrow(cls_train)))
NL_cv   <- matrix(NA_integer_, nrow(cls_train), kmax)
for (f in 1:10) {
  tr <- which(folds_c != f)
  va <- which(folds_c == f)
  m  <- colMeans(cls_train[tr, feats])
  s  <- apply(cls_train[tr, feats], 2, sd)
  NL_cv[va, ] <- nbr_labels(scale(cls_train[tr, feats], center = m, scale = s),
                            scale(cls_train[va, feats], center = m, scale = s),
                            y_tr[tr], kmax)
}
NL_train <- nbr_labels(train_x, train_x, y_tr, kmax)   # in-sample (training) curve
NL_test  <- nbr_labels(train_x, test_x,  y_tr, kmax)   # used once, after k is chosen

tune <- do.call(rbind, lapply(k_grid, function(k) {
  Pcv <- prob_k(NL_cv, k);    cvm <- summ(Pcv, pred_from_P(Pcv), y_tr)
  Ptr <- prob_k(NL_train, k); trm <- summ(Ptr, pred_from_P(Ptr), y_tr)
  data.frame(k = k,
             CV_LogLoss = unname(cvm["LogLoss"]), CV_BalAcc = unname(cvm["BalAcc"]),
             CV_MacroF1 = unname(cvm["MacroF1"]),
             Train_LogLoss = unname(trm["LogLoss"]), Train_BalAcc = unname(trm["BalAcc"]))
}))

k_ll   <- tune$k[which.min(tune$CV_LogLoss)]   # best k by log-loss
k_ba   <- tune$k[which.max(tune$CV_BalAcc)]    # best k by balanced accuracy
k_star <- k_ll                                 # FINAL choice: log-loss
```

We searched odd k from 1 to `r kmax` with 10-fold CV on the training data only. Each candidate was scored with two metrics: balanced accuracy and log-loss. Log-loss is minimized at $k = `r k_ll`$ and balanced accuracy is maximized at $k = `r k_ba`$; the two metrics `r ifelse(k_ll == k_ba, "agreed", "did not agree")`. We chose the final $k^* = `r k_star`$ using **log-loss**, because it is a strictly proper scoring rule that rewards well-calibrated probabilities and is not affected by the arbitrary argmax decision rule, which matters because we compare kNN to Naive Bayes on probability quality as well as on hard classification.

```{r fig.cap = "kNN classification tuning. Top: log-loss (lower is better). Bottom: macro balanced accuracy (higher is better). Red dashed line: best k chosen by CV log-loss. Blue dotted line: k with best CV balanced accuracy.",echo = FALSE}
plot_df <- rbind(
  data.frame(k = tune$k, Metric = "Log-loss",
             Set = "10-fold CV", Value = tune$CV_LogLoss),
  data.frame(k = tune$k, Metric = "Log-loss",
             Set = "Training", Value = tune$Train_LogLoss),
  data.frame(k = tune$k, Metric = "Balanced accuracy",
             Set = "10-fold CV", Value = tune$CV_BalAcc),
  data.frame(k = tune$k, Metric = "Balanced accuracy",
             Set = "Training", Value = tune$Train_BalAcc))

ggplot(plot_df, aes(x = k, y = Value, color = Set)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = k_star, linetype = "dashed", color = "red") +
  geom_vline(xintercept = k_ba, linetype = "dotted", color = "blue") +
  facet_wrap(~ Metric, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("10-fold CV" = "#8B2346", "Training" = "gray50")) +
  labs(x = "k (number of neighbors)", y = NULL, color = NULL) +
  theme_bw(base_size = 12)
```

```{r, echo = FALSE}
top_cls <- head(tune[order(tune$CV_LogLoss), c("k", "CV_LogLoss", "CV_BalAcc", "CV_MacroF1")], 5)
names(top_cls) <- c("k", "CV log-loss", "CV balanced acc.", "CV macro-F1")
tbl(top_cls, paste0("Top five candidate values of k by cross-validated log-loss. Chosen optimum: k = ",
                    k_star, "."))
```

\clearpage

##  Naive Bayes
```{r, echo = FALSE}
H <- table(pos_scale$Position)# Establish count
prior <- prop.table(H)# establish prior
pr <- rbind(Count = H, Prior = round(prior, 4))# binds count and prior to one table
tbl(pr, "Table of Priors and Counts", rn=TRUE)

```
This table shows the catcher and outfield positions being the most populated.


```{r,echo = FALSE}
fitNB <- naiveBayes(Position ~ ., data = data.frame(Position = cls_train$Position, train_x))

# class prediction and its posterior
mapNB <- predict(fitNB, newdata = test_x, type = "class") 
postNB <- predict(fitNB, newdata = test_x, type = "raw")

#data frame of predictions and posteriors
pred_dfNB <- data.frame(
  predicted = mapNB,
  posterior   = round(apply(postNB, 1, max), 4)
)
# Prediction matrix
tabNB <- table(predicted = predict(fitNB, newdata = test_x), actual = cls_test$Position)
tbl(tabNB, "Prediction Matrix of Positions for Naive Bayes", rn=TRUE)
```


```{r,echo = FALSE}
# Confusion matrix 
cmNB <- confusionMatrix(tabNB)
tbl(cmNB$table, "Confusion Matrix of Positions for Naive Bayes", rn=TRUE)


#log loss scores
yNB    <- lv[y_te]
p_trueNB <- postNB[cbind(1:nrow(test_x), match(as.character(yNB), colnames(postNB)))]
loglossNB <- -mean(log(pmax(p_trueNB, 1e-15)))
```


```{r,echo = FALSE}
#null model
pi_cls_train <- prop.table(table(factor(lv[y_tr], levels = lv)))
maj <- names(which.max(pi_train))
loglossnullNB <- -mean(log(pi_cls_train[as.character(yNB)]))
pred_null <- factor(rep(maj, nrow(test_x)), levels = lv)
#null prediction matrix
tab_null <- table(predicted = pred_null, actual = factor(yNB, levels = lv))

```
The prediction matrix for the null model follows the earlier count and prior, showing that Infield was the most populated position in the data set.

```{r,echo = FALSE}
# null confusion matrix
cm_nullNB <- confusionMatrix(tab_null)
tbl(cm_nullNB$table, "Null Model Confusion Matrix of Positions for Naive Bayes", rn=TRUE)

```

```{r, echo = FALSE}
# prints metric scores for comparison between null and prediction
results <- data.frame(
  Model    = c("Naive Bayes", "No-skill"),
  Accuracy = c(cmNB$overall["Accuracy"], cm_nullNB$overall["Accuracy"]),
  BalAcc   = c(mean(cmNB$byClass[, "Balanced Accuracy"]),
               mean(cm_nullNB$byClass[, "Balanced Accuracy"])),       # baseline 0.5
  Kappa    = c(cmNB$overall["Kappa"], cm_nullNB$overall["Kappa"]),      # baseline 0
  LogLoss  = c(loglossNB, loglossnullNB)
)
results
```
comparison between Naive Bayes and Null model

## Comparing KNN vs Naive Bayes Classification

# 5. Discussion and Real World Implications

# 6. Citations and Data Link

Lahman Baseball Database – Society for American Baseball Research. (2025). Retrieved September 29, 2026, from <https://sabr.org/lahman-database/>

The data is linked [here](https://sabr.app.box.com/s/y1prhc795jk8zvmelfd3jq7tl389y6cd). 

# 7. Appendix

## Data cleaning
