---
title: "kNN Regression and kNN and Naive Bayes Classification on Baseball Data"
author: "Sophia Gavasheli and Jacob Carlson"
date: "October 8, 2026"
fontsize: 9pt
classoption: "aspectratio=169"

output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester

header-includes:
  - \definecolor{Green}{HTML}{1B4332}
  - \definecolor{LightGreen}{HTML}{40916C}
  - \definecolor{Cream}{HTML}{F7F3E9}
  - \definecolor{Red}{HTML}{C41E3A}

  - \usecolortheme[named=Green]{structure}

  - \setbeamercolor{block title}{fg=white,bg=Green}
  - \setbeamercolor{block body}{fg=black,bg=Cream}

  - \setbeamercolor{alerted text}{fg=Red}

  - \setbeamertemplate{navigation symbols}{}
---

```{r setup, include=FALSE}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(e1071)
library(caret)
library(FNN)
library(here)
knitr::opts_chunk$set(echo = FALSE)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})

knitr::opts_chunk$set(comment = "") # removes ##

setwd(here()) # sets path to root of RProject

# Part 1A cleaning code
pos <- read.csv("FieldingPost.csv")
batting <- read.csv("batting.csv")

stats <- merge(batting, pos[, c("playerID", "POS")],
               by = "playerID", all.x = TRUE)# add positions

stats_clean <- stats[stats$lgID == "AL" |
                    stats$lgID == "NL", ] # focuses on American and National League

stats_clean <- stats_clean[!is.na(stats_clean$POS), ]  # removes batting data without positions

stats_clean$BA <- ifelse(stats_clean$AB > 0, stats_clean$H /
                           stats_clean$AB, NA)

by_team <- aggregate(cbind(G, AB, R, H, HR, RBI, SB, BB, SO) ~ teamID + yearID,
                      data = stats_clean, FUN = sum) 
# adds Batting Average
by_team$BA <- by_team$H / by_team$AB

# rename positions 
position_names <- c(
  "1B" = "First Base",
  "2B" = "Second Base",
  "3B" = "Third Base",
  "C"  = "Catcher",
  "CF" = "Center Field",
  "LF" = "Left Field",
  "P"  = "Pitcher",
  "RF" = "Right Field",
  "SS" = "Shortstop"
)

by_team_pos <- aggregate(cbind(AB, H, HR, BB, SO, BA, R) ~ teamID + POS, data = stats_clean, FUN = sum)  
by_team_pos$Position <- position_names[by_team_pos$POS]

# loading in code from part B
source("Group17_Exercise1B.R")
```

## Introduction & Research Questions

Our analysis used historical Major League Baseball data from the Lahman Baseball Database

- **Scope:** MLB teams and players in the American League (AL) and National League (NL)

- **Regression RQ:** Predict runs scored by a team by key variables such as at-bats, runs, hits, home runs, stolen bases, walks, and strikeouts, batting average

  - Around 1,500 observations

- **Classification RQ:** Predict a player's position based on batting average, home run average, base on balls average, and strikeout average

  - Around 40,300 observations

## Exploratory Data Analysis (EDA)

**Data Cleaning**

- Derived batting average (BA) = Hits / At-Bats
- Player statistics were aggregated by team and year
- Batting statistics were summarized across player positions
- Positions combined into four categories: catcher, infield, outfield, and pitcher

**Initial Patterns**

- Hits and home runs show a strong positive relationship
- Runs have a relationship with offensive statistics
- Offensive production varies substantially by fielding position

## Runs vs At Bats Aggregated by Team

```{r, echo=FALSE, fig.width=5, fig.height=3}
ggplot(by_team, aes(x = AB, y = R)) +
  geom_point(col = "steelblue") +
  labs(
    x = "At Bats",
    y = "Runs"
  ) +
  theme_minimal()
```

## Home Runs by Position Aggregated by Team

```{r, echo=FALSE, fig.width=5, fig.height=3}
ggplot(by_team_pos, aes(x = Position, y = HR)) +
  geom_boxplot(col = "steelblue") +
  labs(
    x = "Position",
    y = "Home Runs"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## Data Science Methods & Techniques

**kNN Regression**

- Trained on 70% training data scaled with z-scoring
- Optimal $k$ evaluated from $k = 1, \ldots, 51$ with cross-validated RMSE which is more sensitive to large errors
- Model performance evaluated on 30% test data with RMSE, MAE, and $R^2$
- Useful here because baseball statistics may have nonlinear relationships

**kNN and Naive Bayes Classification**

- Predictors scaled based on at bats
- Multi-class targets: used argmax rule to predict class with largest probability
- Optimal $k$ evaluated with cross-validated balanced accuracy and multi-class log loss 
- Model performance evaluated on 30% test data with
  - Log-Loss: penalizes overconfident wrong predictions
  - Brier Score: threshold-free measure of calibration
  - Balanced Accuracy: balances sensitivity and specificity (macro-averaged)
  - Macro-F1: summarizes precision and recall (macro-averaged)

## Results & Key Findings

**Regression**

- Optimal $k$ value was $k^* = 9$
- Model performed better than no-skill model (lower RMSE/MAE, higher $R^2$)
- The model captures about 82% of the variation in runs per game
  - Offensive stats significantly explain runs scored per game

```{r, echo=FALSE}
tbl(results, "kNN regression on the held-out test set versus the null (mean-only) model.")
```

## Results & Key Findings (con't)

**Classification**

- Optimal $k$ valued was chosen based on CV log-loss, $k^* = 51$
- Both kNN and Naive Bayes outperform the no-skill baseline
  - kNN produces slightly better probabilistic predictions (lower Log-Loss and Brier Score)
  - Naive Bayes has slightly higher Balanced Accuracy and Macro-F1 (stronger overall class discrimination)
- Which model is best? kNN performs well for probability quality, while Naive Bayes performs well for balanced classification performance.
- Can batting statistics predict player position?
  - Both models perform better than no-skill baseline, so yes.

```{r, echo=FALSE}
tbl(final_comp, "Test-set comparison of the no-skill baseline, kNN, and Naive Bayes.")
```

## Calibration Plot

```{r, echo=FALSE, fig.width=5, fig.height=3}
ggplot(calib, aes(x = mean_p, y = frac_obs, color = Model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 1) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = c("steelblue", "#C41E3A")) +
  scale_size_continuous(range = c(1.5, 5), name = "Predictions in bin") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean Predicted Probability",
       y = "Observed Fraction of True Class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right", panel.grid.minor = element_blank())
```


## Impact & Discussion

- The analysis demonstrates that relatively simple machine-learning methods can extract meaningful information from baseball statistics
  - Runs scored per game can be predicted from offensive stats
  - Player positions can be predicted from player performance
- The results also demonstrate why multiple evaluation metrics are necessary: kNN performs better on probabilistic metrics, while Naive Bayes performs slightly better on balanced classification metrics.