---
title: "kNN Regression and kNN and Naive Bayes Classification on Baseball Data"
author: "Sophia Gavasheli and Jacob Carlson"
date: "October 8, 2026"
fontsize: 9pt
classoption: "aspectratio=169"

output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester

header-includes:
  - \definecolor{Green}{HTML}{1B4332}
  - \definecolor{LightGreen}{HTML}{40916C}
  - \definecolor{Cream}{HTML}{F7F3E9}
  - \definecolor{Red}{HTML}{C41E3A}

  - \usecolortheme[named=Green]{structure}

  - \setbeamercolor{block title}{fg=white,bg=Green}
  - \setbeamercolor{block body}{fg=black,bg=Cream}

  - \setbeamercolor{alerted text}{fg=Red}

  - \setbeamertemplate{navigation symbols}{}
---

```{r setup, include=FALSE}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(e1071)
library(caret)
library(FNN)
library(here)
knitr::opts_chunk$set(echo = FALSE)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})

knitr::opts_chunk$set(comment = "") # removes ##

setwd(here()) # sets path to root of RProject
```

```{r run-first,ref.label= "Source", include = FALSE}
```
## Introduction & Research Questions

Our analysis used historical Major League Baseball data from the Lahman Baseball Database

- **Scope:** MLB teams and players in the American League (AL) and National League (NL)

- **Regression RQ:** Predict runs scored by a team by key variables such as at-bats, runs, hits, home runs, stolen bases, walks, and strikeouts, batting average

  - Around 1,500 observations

- **Classification RQ:** Predict a player's position based on batting average, home run average, base on balls average, and strikeout average

  - Around 40,300 observations

## Exploratory Data Analysis (EDA)

**Data Cleaning**

- Derived batting average (BA) = Hits / At-Bats
- Player statistics were aggregated by team and year
- Batting statistics were summarized across player positions
- Positions combined into four categories: catcher, infield, outfield, and pitcher

**Initial Patterns**

- Hits and home runs show a strong positive relationship
- Runs have a relationship with offensive statistics
- Offensive production varies substantially by fielding position

## Runs vs At Bats Aggregated by Team

```{r, echo=FALSE, fig.width=5, fig.height=3}
ggplot(team, aes(x = AB, y = R)) +
  geom_point(col = "steelblue") +
  labs(
    x = "At Bats",
    y = "Runs"
  ) +
  theme_minimal()
```

## Home Runs by Position 

```{r, echo=FALSE, fig.width=5, fig.height=3}
ggplot(pos_no_combine, aes(x = Position, y = HR)) +
  geom_boxplot(col = "steelblue") +
  labs(
    x = "Position",
    y = "Home Runs"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## Data Science Methods & Techniques

**kNN Regression**

- Trained on 70% training data scaled with z-scoring
- Optimal $k$ evaluated from $k = 1, \ldots, 51$ with cross-validated RMSE which is more sensitive to large errors
- Model performance evaluated on 30% test data with RMSE, MAE, and $R^2$
- Useful here because baseball statistics may have nonlinear relationships

**kNN and Naive Bayes Classification**

- Predictors scaled based on at bats
- Multi-class targets: used argmax rule to predict class with largest probability
- Optimal $k$ evaluated with cross-validated balanced accuracy and multi-class log loss 
- Model performance evaluated on 30% test data with
  - Log-Loss: penalizes overconfident wrong predictions
  - Brier Score: threshold-free measure of calibration
  - Balanced Accuracy: balances sensitivity and specificity (macro-averaged)
  - Macro-F1: summarizes precision and recall (macro-averaged)

## Results & Key Findings

**Regression**

- Optimal $k$ value was $k^* = 9$
- Model performed better than no-skill model (lower RMSE/MAE, higher $R^2$)
- The model captures about 82% of the variation in runs per game
  - Offensive stats significantly explain runs scored per game

```{r, echo=FALSE}
tbl(results, "kNN regression on the held-out test set versus the null (mean-only) model.")
```

## Results & Key Findings (con't)

**Classification**

- Optimal $k$ valued was chosen based on CV log-loss, $k^* = 51$
- Both kNN and Naive Bayes outperform the no-skill baseline
  - kNN produces slightly better probabilistic predictions (lower Log-Loss and Brier Score)
  - Naive Bayes has slightly higher Balanced Accuracy and Macro-F1 (stronger overall class discrimination)
- Which model is best? kNN performs well for probability quality, while Naive Bayes performs well for balanced classification performance.
- Can batting statistics predict player position?
  - Both models perform better than no-skill baseline, so yes.

```{r, echo=FALSE}
tbl(final_comp, "Test-set comparison of the no-skill baseline, kNN, and Naive Bayes.")
```

## Calibration Plot

```{r, echo=FALSE, fig.width=5, fig.height=3}
ggplot(calib, aes(x = mean_p, y = frac_obs, color = Model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 1) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = c("steelblue", "#C41E3A")) +
  scale_size_continuous(range = c(1.5, 5), name = "Predictions in bin") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean Predicted Probability",
       y = "Observed Fraction of True Class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right", panel.grid.minor = element_blank())
```


## Impact & Discussion

- The analysis demonstrates that relatively simple machine-learning methods can extract meaningful information from baseball statistics
  - Runs scored per game can be predicted from offensive stats
  - Player positions can be predicted from player performance
- The results also demonstrate why multiple evaluation metrics are necessary: kNN performs better on probabilistic metrics, while Naive Bayes performs slightly better on balanced classification metrics.

```{r Source, include = FALSE}
## ----setup, include=FALSE-----------------------------------------------------------------------

## ----echo = FALSE---------------------------------------------------------------------------------------
pos     <- read.csv("Fielding.csv")
batting <- read.csv("batting.csv")
team    <- read.csv("Teams.csv")
# Getting highest position played for players
primary_pos <- pos %>%
  group_by(playerID, yearID, POS) %>%
  summarise(G = sum(G), .groups = "drop") %>%
  group_by(playerID, yearID) %>%
  slice_max(G, n = 1, with_ties = FALSE) %>%
  select(playerID, yearID, POS)

# Merge Position into Player and Year to condense data
stats <- merge(batting, primary_pos,
               by = c("playerID", "yearID"), all.x = TRUE)

# Keeps only American and National League
stats_clean <- stats[stats$lgID %in% c("AL", "NL"), ]
stats_clean <- stats_clean[!is.na(stats_clean$POS), ]


####### REGRESSION 
## Separates by team for regression prediction. 
team <- team[complete.cases(team[, c("AB", "R")]), ] 
team_clean <- team[team$lgID %in% c("AL","NL"),] # Keeps only AL and NL
team_clean$X1B <- team_clean$H - team_clean$X2B - team_clean$X3B - team_clean$HR # Creates X1B
team_clean$HBPxBB <- team_clean$BB + team_clean$HBP # Creates HBPxBB


# Takes only the necessary data
team_clean <- team_clean[,c("yearID", "teamID","G", "R", "X1B", "X2B", "X3B", 
                            "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")] 

#Divides stats by games to establish per game base
stats <- c("R", "X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB")
team_clean[stats] <- team_clean[stats] / team_clean$G

team_clean <- team_clean[!is.na(team_clean$SF), ]# Removes NaN values from SF
team_clean <- team_clean[!is.na(team_clean$HBPxBB), ]# Removes NaN values from HBPxBB

# List of predictors for training
num_vars <- c("X1B", "X2B", "X3B", "HR", "SO", "HBPxBB", "SB", "SF", "CS", "AB") 


##### CLASSIFICATION AND BAYES
# Keeps only At Bats > 100
pos_data <- stats_clean[stats_clean$AB >= 100, ] 

##################################
## don't combine positions for boxplot
pos_no_combine <- pos_data
pos_names <- c(
  "1B" = "First Base",
  "2B" = "Second Base",
  "3B" = "Third Base",
  "C"  = "Catcher",
  "CF" = "Center Field",
  "LF" = "Left Field",
  "P"  = "Pitcher",
  "RF" = "Right Field",
  "SS" = "Shortstop"
)
pos_no_combine$Position <- factor(pos_names[pos_no_combine$POS])
pos_no_combine <- pos_no_combine[, c("playerID", "yearID", "Position", "HR")]

###########################
# Changes The infield positions into a common infield tab
for(i in 1:length(pos_data$POS)){
  if(pos_data$POS[i] == "1B" || 
     pos_data$POS[i] == "2B" ||
     pos_data$POS[i] == "3B" ||
     pos_data$POS[i] == "SS" ){
    pos_data$POS[i] = "IF"
  }
}
# For easier referral
position_names <- c(
  "IF" = "Infield",
  "C"  = "Catcher",
  "OF" = "Outfield",
  "P"  = "Pitcher"
)

#Factors Positions into levels
pos_data$Position <- factor(position_names[pos_data$POS])

#Keep only necessary data
pos_data <- pos_data[, c("playerID", "yearID", "Position",
                         "H", "HR", "BB", "SO", "SB", "AB")]


# creates scaled pos_data for Classification comparison and naive Bayes
pos_scale <- pos_data

pos_scale$BA <- pos_data$H  / pos_data$AB
pos_scale$HRA <- pos_data$HR / pos_data$AB
pos_scale$BBA <- pos_data$BB / pos_data$AB
pos_scale$SOA <- pos_data$SO / pos_data$AB

#Adds scaled variables into pos_scaled
pos_scale <- pos_scale[, c("playerID", "yearID", "Position",
                                  "BA", "HRA", "BBA", "SOA")]
# List of Predictors for pos_scaled
feats <- c("BA", "HRA", "BBA", "SOA") 


## ----echo = FALSE---------------------------------------------------------------------------------------
# table formatting function
tbl <- function(x, cap, digits = 3, rn = FALSE, ...) {
  knitr::kable(x, booktabs = TRUE, caption = cap, digits = digits,
               position = "H", row.names = rn, ...)
}

d2  <- unlist((team_clean[1, num_vars] - team_clean[2, num_vars])^2)
z   <- scale(team_clean[, num_vars])
d2z <- (z[1, ] - z[2, ])^2
demo <- rbind("Raw (share of squared distance, pct)"          = 100 * d2  / sum(d2),
              "Z-scored (share of squared distance, pct)"     = 100 * d2z / sum(d2z))
tbl(demo, "Share of squared Euclidean distance contributed by each predictor (first two games) in the kNN Regression Dataset.",
    digits = 1, rn = TRUE)


## ----echo = FALSE---------------------------------------------------------------------------------------
d2  <- unlist((pos_scale[1, feats] - pos_scale[2, feats])^2)
z   <- scale(pos_scale[, feats])
d2z <- (z[1, ] - z[2, ])^2
demo <- rbind("Raw (share of squared distance, pct)"          = 100 * d2  / sum(d2),
              "Z-scored (share of squared distance, pct)"     = 100 * d2z / sum(d2z))
tbl(demo, "Share of squared Euclidean distance contributed by each predictor (first two team-seasons) in the Classification Dataset.",
    digits = 1, rn = TRUE)


## ----echo = FALSE---------------------------------------------------------------------------------------
#regression split 70/30
set.seed(4654)
idx_reg   <- sample(nrow(team_clean), round(0.7 * nrow(team_clean)))
train_raw <- team_clean[idx_reg, num_vars]
test_raw  <- team_clean[-idx_reg, num_vars]
train_y   <- team_clean$R[idx_reg]
test_y    <- team_clean$R[-idx_reg]

mu_reg      <- colMeans(train_raw)
sd_reg      <- apply(train_raw, 2, sd)
training_df <- scale(train_raw, center = mu_reg, scale = sd_reg)
testing_df  <- scale(test_raw,  center = mu_reg, scale = sd_reg)

# classification split (knn and nb)

set.seed(4654)
idx_cls <- unlist(lapply(split(seq_len(nrow(pos_scale)), pos_scale$Position),
                         function(i) sample(i, round(0.7 * length(i)))))
cls_train <- pos_scale[idx_cls, ]
cls_test  <- pos_scale[-idx_cls, ]

mu_cls  <- colMeans(cls_train[, feats])
sd_cls  <- apply(cls_train[, feats], 2, sd)
train_x <- scale(cls_train[, feats], center = mu_cls, scale = sd_cls)
test_x  <- scale(cls_test[, feats],  center = mu_cls, scale = sd_cls)

lv   <- levels(pos_scale$Position)
K    <- length(lv)
y_tr <- as.integer(cls_train$Position)
y_te <- as.integer(cls_test$Position)

# null entropy log loss baseline for knn and nb classification
pi_train     <- prop.table(table(cls_train$Position))[lv]
null_entropy <- -sum(pi_train * log(pi_train))


## ----echo=FALSE, fig.cap = "kNN regression tuning: cross-validated and training RMSE versus k. The red point marks the CV-optimal k.",echo = FALSE----
set.seed(4654)
folds <- sample(rep(1:10, length.out = nrow(train_raw)))

# creates folds, scaling inside folds
fold_data <- lapply(1:10, function(f) {
  tr <- train_raw[folds != f, ] #training fold
  va <- train_raw[folds == f, ] #validation fold
  m  <- colMeans(tr)
  s  <- apply(tr, 2, sd)
  list(tr_x = scale(tr, center = m, scale = s), #response
       va_x = scale(va, center = m, scale = s),
       tr_y = train_y[folds != f], #predictor
       va_y = train_y[folds == f])
})

# function to train knn on inputted val of k on the 10 folds created with fold_data()
cv_rmse <- function(k) {
  sq_err <- unlist(lapply(fold_data, function(d) {
    pred <- knn.reg(train = d$tr_x, test = d$va_x, y = d$tr_y, k = k)$pred
    (d$va_y - pred)^2
  }))
  sqrt(mean(sq_err))
}

# train k without CV
train_rmse <- function(k) {
  pred <- knn.reg(train = training_df, test = training_df, y = train_y, k = k)$pred
  sqrt(mean((train_y - pred)^2))
}

k_cv       <- 1:51
cv_results <- data.frame(k = k_cv,
                         RMSE = sapply(k_cv, cv_rmse),
                         Train_RMSE = sapply(k_cv, train_rmse))
k_best <- cv_results$k[which.min(cv_results$RMSE)]

ggplot(cv_results, aes(x = k)) +
  geom_line(aes(y = RMSE, color = "10-fold CV"), linewidth = 1) +
  geom_line(aes(y = Train_RMSE, color = "Training"), linewidth = 1) +
  geom_point(data = cv_results[cv_results$k == k_best, ],
             aes(y = RMSE), color = "red", size = 4) +
  scale_color_manual(values = c("10-fold CV" = "#8B2346", "Training" = "orange")) +
  labs(
       x = "Number of Neighbors (k)", y = "RMSE of Runs per game", color = NULL) +
  theme_bw(base_size = 12)


## ----echo = FALSE---------------------------------------------------------------------------------------
top_k <- head(cv_results[order(cv_results$RMSE), c("k", "RMSE", "Train_RMSE")], 5)
names(top_k) <- c("k", "CV RMSE", "Training RMSE")
tbl(top_k, paste0("Top five candidate values of k by cross-validated RMSE. The chosen optimum is k* = ",
                  k_best, "."))


## ----echo=FALSE-----------------------------------------------------------------------------------------
# final model
final_pred <- knn.reg(train = training_df, test = testing_df, y = train_y, k = k_best)$pred

rmse      <- function(a, b) sqrt(mean((a - b)^2))
mae       <- function(a, b) mean(abs(a - b))
r_squared <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)

null_prediction <- rep(mean(train_y), length(test_y))

results <- data.frame(
  Model = c("Null (training mean)", paste0("kNN (k* = ", k_best, ")")),
  RMSE  = c(rmse(test_y, null_prediction), rmse(test_y, final_pred)),
  MAE   = c(mae(test_y, null_prediction),  mae(test_y, final_pred)),
  R2    = c(r_squared(test_y, null_prediction), r_squared(test_y, final_pred))
)

results[, -1] <- round(results[, -1], 3)

names(results)[4] <- "R-squared"
tbl(results, "kNN regression on the held-out test set versus the null (mean-only) model.")


## ----echo=FALSE-----------------------------------------------------------------------------------------
# helper to convert classes with 1 hot encoding
onehot   <- function(y_int) diag(K)[y_int, , drop = FALSE]
# calc log loss
logloss  <- function(P, y_int) -mean(log(pmax(P[cbind(seq_along(y_int), y_int)], 1e-15)))
# calc brier score
brier    <- function(P, y_int) mean(rowSums((P - onehot(y_int))^2))
#argmax
pred_from_P <- function(P) max.col(P, ties.method = "first")

# create confusion matrix
conf_tab <- function(pred_int, y_int) {
  table(Predicted = factor(pred_int, levels = 1:K, labels = lv),
        Actual    = factor(y_int,    levels = 1:K, labels = lv))
}

# metrics per class, input confusion matrix
per_class <- function(cm) {
  TP <- diag(cm); FP <- rowSums(cm) - TP; FN <- colSums(cm) - TP
  TN <- sum(cm) - TP - FP - FN
  sens <- TP / (TP + FN); spec <- TN / (TN + FP)
  data.frame(
    Sensitivity = sens,
    Specificity = spec,
    F1          = 2 * TP / (2 * TP + FP + FN),
    BalAcc      = (sens + spec) / 2
  )
}

# summary to evaluate model performance
summ_metrics <- function(P, y_int) {
  cm  <- conf_tab(pred_from_P(P), y_int)
  pc  <- per_class(cm)
  b   <- brier(P, y_int)
  
  c(LogLoss = logloss(P, y_int),
    Brier   = b,
    RMSE    = sqrt(b),
    BalAcc  = mean(pc$BalAcc, na.rm = TRUE),
    MacroF1 = mean(pc$F1, na.rm = TRUE))
}

# knn grid
k_grid <- seq(1, 51, by = 2)
kmax   <- max(k_grid)

# neighbor labeks
nbr_labels <- function(train_mat, query_mat, y_int, kmax) {
  nn <- FNN::knnx.index(train_mat, query_mat, k = kmax)
  matrix(y_int[nn], nrow = nrow(nn))
}

# smoothed vote fractions
prob_k <- function(NL, k) {
  votes <- sapply(seq_len(K), function(cl) rowSums(NL[, seq_len(k), drop = FALSE] == cl))
  P <- (votes + 1) / (k + K)
  colnames(P) <- lv
  P
}

# 10-fold CV loop
set.seed(4654)
folds_c <- sample(rep(1:10, length.out = nrow(cls_train)))
NL_cv   <- matrix(NA_integer_, nrow(cls_train), kmax)

for (f in 1:10) {
  tr <- which(folds_c != f)
  va <- which(folds_c == f)
  m  <- colMeans(cls_train[tr, feats])
  s  <- apply(cls_train[tr, feats], 2, sd)
  NL_cv[va, ] <- nbr_labels(
    scale(cls_train[tr, feats], center = m, scale = s),
    scale(cls_train[va, feats], center = m, scale = s),
    y_tr[tr], kmax
  )
}

NL_train <- nbr_labels(train_x, train_x, y_tr, kmax)

# tuning
tune <- do.call(rbind, lapply(k_grid, function(k) {
  Pcv <- prob_k(NL_cv, k);     cvm <- summ_metrics(Pcv, y_tr)
  Ptr <- prob_k(NL_train, k);  trm <- summ_metrics(Ptr, y_tr)
  
  data.frame(
    k             = k,
    CV_LogLoss    = unname(cvm["LogLoss"]),
    Train_LogLoss = unname(trm["LogLoss"]),
    CV_BalAcc     = unname(cvm["BalAcc"]),
    Train_BalAcc  = unname(trm["BalAcc"]),
    CV_Brier      = unname(cvm["Brier"]),
    CV_RMSE       = unname(cvm["RMSE"]),
    CV_MacroF1    = unname(cvm["MacroF1"])
  )
}))

# optimal k choices based on the two evaluation metrics
k_ll <- tune$k[which.min(tune$CV_LogLoss)]
k_ba <- tune$k[which.max(tune$CV_BalAcc)]
k_star = k_ll #final choice is log loss

top_k_cv <- head(tune[order(tune$CV_LogLoss), c("k", "CV_LogLoss", "CV_BalAcc")], 5)
top_k_tr <- head(tune[order(tune$Train_LogLoss), c("k", "Train_LogLoss", "Train_BalAcc")], 5)

names(top_k_cv) <- c("k", "CV log-loss", "CV bal. acc.")
names(top_k_tr) <- c("k", "Train log-loss", "Train bal. acc.")

tbl(top_k_tr, "Top five candidate values of k by training log-loss and balanced accuracy.")
tbl(top_k_cv, "Top five candidate values of k by cross-validated log-loss and balanced accuracy.")


## ----echo=FALSE, fig.cap="kNN classification tuning: cross-validated and training balanced accuracy and log-loss vs k. The dots mark the optimal k."----

# reshape tune data into long format for faceting
tune_long <- data.frame(
  k      = rep(tune$k, 4),
  Metric = rep(c("Log-Loss", "Log-Loss", "Balanced Accuracy", "Balanced Accuracy"), each = nrow(tune)),
  Set    = rep(c("10-fold CV", "Train", "10-fold CV", "Train"), each = nrow(tune)),
  Value  = c(tune$CV_LogLoss, tune$Train_LogLoss, tune$CV_BalAcc, tune$Train_BalAcc)
)

# optimal k values
opt_points <- data.frame(
  k      = c(k_ll, k_ll, k_ba, k_ba),
  Metric = c("Log-Loss", "Log-Loss", "Balanced Accuracy", "Balanced Accuracy"),
  Set    = c("10-fold CV", "Train", "10-fold CV", "Train"),
  Value  = c(
    tune$CV_LogLoss[tune$k == k_ll],
    tune$Train_LogLoss[tune$k == k_ll],
    tune$CV_BalAcc[tune$k == k_ba],
    tune$Train_BalAcc[tune$k == k_ba]
  )
)

# plot
ggplot(tune_long, aes(x = k, y = Value, color = Set)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ Metric, scales = "free_y", nrow = 1) +
geom_point(data = opt_points, aes(x = k, y = Value, color = Set), size = 4, inherit.aes = FALSE) +
  scale_color_manual(values = c("10-fold CV" = "#8B2346", "Train" = "orange")) +
  labs(
    x = "Number of Neighbors (k)",
    y = "Metric Value",
    color = "Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )


## ----echo=FALSE-----------------------------------------------------------------------------------------
# Generate neighbor labels for the test set using kmax, then subset for k_star
NL_test <- nbr_labels(train_x, test_x, y_tr, kmax)
P_test  <- prob_k(NL_test, k_star)

# Compute test metrics
test_metrics <- summ_metrics(P_test, y_te)

# relative improvement
rel_improvement <- 1 - (test_metrics["LogLoss"] / null_entropy)

# confusion matrix
test_pred <- pred_from_P(P_test)
test_cm <- conf_tab(test_pred, y_te)
tbl(as.data.frame.matrix(test_cm), 
    paste0("Confusion matrix for the final kNN test set evaluation (k* = ", k_star, ")."), rn=TRUE)

# comparison table
comparison_df <- data.frame(
  Metric = c("Log-Loss", "Relative Improvement over Null", "Brier Score", "Balanced Accuracy (Macro)", "Macro-F1"),
  No_Skill_Baseline = c(round(null_entropy, 4), "0.00%", 1 - sum(pi_train^2), "0.5", mean(2 * pi_train / (1 + pi_train))),
  Final_Model_k_star = c(
    round(test_metrics["LogLoss"], 4),
    paste0(round(rel_improvement * 100, 2), "%"),
    round(test_metrics["Brier"], 4),
    round(test_metrics["BalAcc"], 4),
    round(test_metrics["MacroF1"], 4)
  )
)
names(comparison_df) <- c("Metric", "No-Skill Baseline", paste0("kNN (k* = ", k_star, ")"))
tbl(comparison_df, paste0("Performance comparison of final kNN model (k* = ", k_star, ") against the no-skill baseline on the test set."))


## ----echo = FALSE---------------------------------------------------------------------------------------
train_counts <- as.integer(table(cls_train$Position)[lv])
pi_tab <- rbind(Count      = train_counts,
                Proportion = round(as.numeric(pi_train), 4))
colnames(pi_tab) <- lv

tbl(pi_tab, "Training-set class counts and proportions ($\\pi_k$).",
    rn = TRUE)


## ----echo = FALSE---------------------------------------------------------------------------------------
fitNB <- naiveBayes(Position ~ ., data = data.frame(Position = cls_train$Position, train_x))

# test-set class probabilities
P_nb <- predict(fitNB, newdata = as.data.frame(test_x), type = "raw")[, lv, drop = FALSE]

# confusion matrix 
nb_metrics <- summ_metrics(P_nb, y_te)
nb_cm      <- conf_tab(pred_from_P(P_nb), y_te)

tbl(as.data.frame.matrix(nb_cm),
    "Confusion matrix for Naive Bayes on the test set.", rn = TRUE)


## ----echo = FALSE---------------------------------------------------------------------------------------
# no-skill baselines from the training priors
pi_nb      <- as.numeric(prop.table(table(factor(lv[y_tr], levels = lv))))
entropy_nb <- -sum(pi_nb * log(pi_nb))

null_base <- c(
  LogLoss = entropy_nb,
  Brier   = 1 - sum(pi_nb^2),
  BalAcc  = 0.5,
  MacroF1 = mean(2 * pi_nb / (1 + pi_nb))
)

# null model confusion matrix
maj_int <- which.max(table(factor(y_tr, levels = 1:K)))
null_pred <- rep(maj_int, length(y_te))
null_cm   <- conf_tab(null_pred, y_te)

tbl(as.data.frame.matrix(null_cm),
    "Confusion matrix for the no-skill model on the test set (always predicts the majority class).",
    rn = TRUE)

# relative log-loss improvement over no-skill
rel_improvement_nb <- 1 - nb_metrics["LogLoss"] / null_base["LogLoss"]

comparison_nb <- data.frame(
  Metric = c("Log-Loss", "Relative Improvement over Null", "Brier Score",
             "Balanced Accuracy (Macro)", "Macro-F1"),
  NoSkill = c(round(null_base["LogLoss"], 4),
              "0.00%",
              round(null_base["Brier"], 4),
              round(null_base["BalAcc"], 4),
              round(null_base["MacroF1"], 4)),
  NB = c(round(nb_metrics["LogLoss"], 4),
         paste0(round(rel_improvement_nb * 100, 2), "%"),
         round(nb_metrics["Brier"], 4),
         round(nb_metrics["BalAcc"], 4),
         round(nb_metrics["MacroF1"], 4))
)
names(comparison_nb) <- c("Metric", "No-Skill Baseline", "Naive Bayes")
tbl(comparison_nb, "Performance comparison of Naive Bayes against the no-skill baseline on the test set.")


## ----echo = FALSE---------------------------------------------------------------------------------------
final_comp <- data.frame(
  Metric  = c("Log-Loss", "Brier Score", "Balanced Accuracy (Macro)", "Macro-F1"),
  NoSkill = round(null_base[c("LogLoss", "Brier", "BalAcc", "MacroF1")], 4),
  kNN     = round(test_metrics[c("LogLoss", "Brier", "BalAcc", "MacroF1")], 4),
  NB      = round(nb_metrics[c("LogLoss", "Brier", "BalAcc", "MacroF1")], 4)
)
names(final_comp) <- c("Metric", "No-Skill", paste0("kNN (k* = ", k_star, ")"), "Naive Bayes")
tbl(final_comp, "Test-set comparison of the no-skill baseline, kNN, and Naive Bayes.")


## ----echo=FALSE, fig.cap="Calibration plot comparing kNN and Naive Bayes Classification. Diagonal line represents perfect calibration.", fig.pos='H'----

calib_df <- function(P, y_int, model, bins = 10) {
  df <- data.frame(p   = as.vector(P),             # every class-probability
                   obs = as.vector(onehot(y_int))) # 1 if that class was the true one
  df$bin <- cut(df$p, breaks = seq(0, 1, length.out = bins + 1), include.lowest = TRUE)
  out <- do.call(rbind, lapply(split(df, droplevels(df$bin)), function(d) {
    data.frame(mean_p = mean(d$p), frac_obs = mean(d$obs), n = nrow(d))
  }))
  out$Model <- model
  out
}

calib <- rbind(
  calib_df(P_test, y_te, paste0("kNN (k* = ", k_star, ")")),
  calib_df(P_nb,   y_te, "Naive Bayes")
)

ggplot(calib, aes(x = mean_p, y = frac_obs, color = Model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 1) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = c("steelblue", "lightcoral")) +
  scale_size_continuous(range = c(1.5, 5), name = "Predictions in bin") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean predicted probability",
       y = "Observed fraction of true class") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())


```